---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align='center'
)
```

# ConSpec

R package for contrastive spectral analysis.

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

### Authors

Yunhui Qi, Peng Liu, Yumou Qiu

### Contact

yunhuiqistat@gmail.com (Yunhui Qi)

## Installation

You can install the development version of ConSpec from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("yunhuiqistat/ConSpec")
```

## Website

Check the vignette at https://yunhuiqistat.github.io/ConSpec/.


## Introduction

The goal of CEA is to conduct contrastive eigen analysis including contrastive principal component (cPCA) and its sparse version (scPCA), contrastive cross-covariance analysis (3CA) and its sparse version (3CA). (s)cPCA can be implemented by function `cPCA`. (s)3CA can be implemented by function `cCCA`.


```{r load libraries for this intro, message=FALSE}
library(ConSpec)
library(ggpubr)
library(ggplot2)
library(ggsci)
```


### cPCA 

`eta_tuning_general()`

* Input

  + `Xt` data frame for target group, samples in rows, variables in columns. If PCA is applied to correlation matrix, Xt should be scaled and centered.  
  + `Xa` data frame for ancillary group, samples in rows, variables in columns. If cPCA is applied to correlation matrix, Xc should also be scaled and centered.  
  + `eta_range` a vector containing candidate eta range to be tuning from.  
  + `plot` a logical quantity indicating whether plot the ARI(nuisance) and silhouette score.  
  + `sparse_trt` parameter controlling the sparsity of target sparse PCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_ctst` parameter controlling the sparsity of scPCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `num_comps` if scPCA, specify the number of contrastive components that will be estimated by PMD, default is 20.
  + `km_cluster` number showing number of nuisance clusters defined from kmeans clustering on target PCA. 
  + `ckm_cluster` number of interesting clusters defined from kmeans clustering on cPCA/scPCA. 


* Output

  + A list containing selected value of $\eta$  `eta_opt`, the vector of ARI(nuisance) `nuisance_ARI` and the vector of Silhouette `silhouette`, and the ggplot object `p`.


`cPCA()`

* Input

  + `Xt` data frame for target group, samples in rows, variables in columns.
  + `Xa` data frame for ancillary group, samples in rows, variables in columns.
  + `eta` tuning parameter controlling how much ancillary variation should be contrasted off from the target variation. It can be theo `eta_opt` from function eta_tuning(). 
  + `sparse_ctst` parameter controlling the sparsity of scPCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_trt` parameter controlling the sparsity of target sparse PCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_ctrl` parameter controlling the sparsity of ancillary sparse PCA, if NULL, use cPCA, else use scPCA. For use of scPCA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.


* Output

  + A list containing contrastive scores `cPC` and loadings `cU`, target scores `PCt` and loadings `Ut`, and ancillary scores `PCa` and loadings `Ua`.

1. We load the simulated data from the `data(intro)`, and use `eta_tuning()` to get optimal value of tuning parameter $\eta$. 


```{r eta-tuning-for-cPCA}
eta_range <- seq(0.1, 5, 0.05)
eta_opt <- eta_tuning_general(Xt = intro$cPCA$Xt, Xa = intro$cPCA$Xa, eta_range = eta_range, plot = TRUE, sparse_trt = NULL, sparse_ctst = NULL, num_comps = 20, km_cluster = 2, ckm_cluster = 2)$eta_opt
```

2. We then input the data and the resulting $\eta$ value to cPCA.

```{r cPCA-usage}
# speccify sparsity parameters to be NULL for cPCA
cpca_res <- cPCA(Xt = intro$cPCA$Xt, Xa = intro$cPCA$Xa, eta = eta_opt, sparse_ctst = NULL, sparse_trt = NULL, sparse_ctrl = NULL)
```

3. We get the score plot of cPCA. In comparison, we also get the score plot of target PCA. Here, as in the paper, we have nuisance subgroup factor $C$ and interesting subgroup factor $B$. In the score plot, we see the first target score is associated with the nuisance subgroups, while the first contrastive score is associated with the interesting subgroups. This validates the removal of nuisance effect and discovery of hidden subgroups by cPCA.

```{r cPCA-usage-score-plot, fig.height=5, fig.width=10}
plot_data <- data.frame(cPC1 = cpca_res$cPC[,1], cPC2 = cpca_res$cPC[,2],
                        PC1 = cpca_res$PCt[,1], PC2 = cpca_res$PCt[,2],
                        intro$cPCA$covariates_t)
cP <- ggplot(plot_data)+
        geom_point(aes(x = cPC1, y = cPC2, color = B, shape = C))+
        labs(title = "Contrastive", color = "Interesting \n subgroups", shape = "Nuisance \nsubgroups")+
        theme_bw()+
        theme(panel.grid = element_blank(),
              text = element_text(size = 14),  # Adjust text size
              axis.title = element_text(size = 16),  # Adjust axis title size
              axis.text = element_text(size = 12))
Pt <- ggplot(plot_data)+
        geom_point(aes(x=PC1, y=PC2, color= B, shape = C), show.legend = FALSE)+
        labs(title = "Target", color = "Interesting \n subgroups", shape = "Nuisance \nsubgroups")+
        theme_bw()+
        theme(panel.grid = element_blank(),
              text = element_text(size = 14),  # Adjust text size
              axis.title = element_text(size = 16),  # Adjust axis title size
              axis.text = element_text(size = 12))
p_score <- ggarrange(Pt,cP, nrow = 1, common.legend = TRUE, legend = "bottom", align = "v")
print(p_score)
```

### scPCA

To illustrate the use of function `cPCA()` for sparse cPCA, we use the barley expression data example. For details of data processing and references, please refer to `data(barley)`.

1. We load the processed data from the `data(intro)`, and use assigned $\eta = 1$, we also specify the sparsity parameters to be 0.3.

```{r scPCA-usage}
set.seed(111)
# specify the sparsity parameters for scPCA
scPCA_res <- cPCA(Xt = intro$scPCA$Xt, Xa = intro$scPCA$Xa, eta = 1, sparse_ctst = 0.3, sparse_trt = 0.3, sparse_ctrl = 0.3)
```


2. We get the score plot of scPCA. In comparison, we also get the score plot of target PCA. As seen from the figure, Genotype effect is removed by scPCA, and the fungus subgroups is revealed by the first scPCA score.

```{r scPCA-usage-score-plot, fig.height=5, fig.width=10}
PCt <- scPCA_res$PCt
PCa <- scPCA_res$PCa
cPC <- scPCA_res$cPC
Ut <- scPCA_res$Ut
Ua <- scPCA_res$Ua
cU <- scPCA_res$cU

plot_data_t <- data.frame(cPC1 = cPC[,1], cPC2 = cPC[,2],
                          PC1 = PCt[,1], PC2 = PCt[,2],
                          Genotype = intro$scPCA$covariates_t$Genotype,
                          Isolate = intro$scPCA$covariates_t$Isolate,
                          Time_bi = factor(intro$scPCA$covariates_t$Time_bi, levels = c("Early","Late")))
plot_data_a <- data.frame(PC1 = PCa[,1], PC2 = PCa[,2], 
                          Genotype = intro$scPCA$covariates_a$Genotype,
                          Isolate = intro$scPCA$covariates_a$Isolate,
                          Time = intro$scPCA$covariates_a$Time,
                          Time_bi = factor(intro$scPCA$covariates_a$Time_bi, levels = c("Early","Late")))

cP <- ggplot(plot_data_t)+
  geom_point(aes(x = cPC1, y = cPC2, color = Isolate, shape = Genotype), size = 2)+
  theme_bw()+
  labs(title = "Contrastive", color = "Fungus")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()
Pt <- ggplot(plot_data_t)+
  geom_point(aes(x=PC1, y=PC2, color= Isolate, shape = Genotype ), size = 2)+
  labs(title = "Late", color = "Fungus")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()
Pa <- ggplot(plot_data_a)+
  geom_point(aes(x=PC1, y=PC2, color= Isolate, shape = Genotype), size = 2)+
  labs(title = "Early", color = "Fungus")+
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "right")+
  scale_color_d3()

p_score <- ggarrange(Pa, Pt, cP,  nrow=1, ncol=3,  common.legend = TRUE, legend = "bottom")
print(p_score)
```

### 3CA

`cCCA()`

* Input 

  + `Xt` data frame for one data type in target group, samples in rows, variables in columns. To conduct analysis based on cross corrlation matrix, each feature should be centered and standardized.    
  + `Xa` data frame for one data type in ancillary group, samples in rows, variables in columns. To conduct analysis based on cross corrlation matrix, each feature should be centered and standardized.  
  + `Yt` data frame for another data type in target group, samples in rows, variables in columns. To conduct analysis based on cross corrlation matrix, each feature should be centered and standardized.  
  + `Ya` data frame for another data type in ancillary group, samples in rows, variables in columns. To conduct analysis based on cross corrlation matrix, each feature should be centered and standardized.  
  + `eta` tuning parameter controlling how much ancillary variation should be contrasted off from the target variation. If NULL, the estimated eta introduced in paper will be used.
  + `sparse_ctst` parameter controlling the sparsity of s3CA, if NULL, use 3CA, else use s3CA. For use of s3CA, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_trt` parameter controlling the sparsity of target sparse cross covariance/correlation analysis, if NULL, use cross covariance/correlation analysis, else use sparse cross covariance/correlation analysis. For use of sparse version, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  
  + `sparse_ctrl` parameter controlling the sparsity of ancillary sparse cross covariance/correlation analysis, if NULL, use cross covariance/correlation analysis, else use sparse cross covariance/correlation analysis. For use of sparse version, if vector, refer sumabss in PMD.cv from package PMA, if single number, refer sumabs in PMD in package PMA.  

* Output

  + A list containing contrastive SVD object `CCA_ctst` including singular values `d`, loadings for X `u`, loadings for Y `v`; target only SVD object `CCA_trt` and ancillary only SVD object `CCA_ctrl`.



1. We load the simulated data from the `data(intro)`, and get the first pair of score from 3CA, then we cluster the scores using kmeans clustreing with two clusters. We would like to check if these scores recover the target specific clusters. 


```{r 3CA}
res_3CA <- cCCA(Xa = intro$cCCA$Xa, Ya = intro$cCCA$Ya, 
                Xt = intro$cCCA$Xt, Yt = intro$cCCA$Yt, eta  = NULL,
                sparse_ctst = NULL, sparse_trt = NULL, sparse_ctrl = NULL)
```

```{r 3CA-usage}
df_score <- data.frame(CCcX=intro$cCCA$Xt %*% res_3CA$CCA_ctst$u[,1], CCcY=intro$cCCA$Yt %*% res_3CA$CCA_ctst$v[,1], labels = intro$cCCA$labels)
clustering <- spec_clust(df_score[,1:2], k = 2, group = intro$cCCA$labels)
df_score$Clusters <- clustering$cluster
ggplot(df_score)+geom_point(aes(x = CCcX, y = CCcY, color= factor(labels), shape = factor(Clusters)), size = 2)+
  labs(title = "Contrastive", color = "True target specific clusters", shape = "Estimated clusters from contrastive")+
  theme_bw()+
  guides(
    shape = guide_legend(
      nrow = 1,
      byrow = TRUE,
      order = 1
    ),
    color = guide_legend(
      nrow = 1,
      byrow = TRUE,
      order = 2
    )
  ) +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 14),  # Adjust text size
    axis.title = element_text(size = 16),  # Adjust axis title size
    axis.text = element_text(size = 12),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.box.just = "center"
  )
cat("ARI for target specific clusters:", clustering$ari, "\n")
```


### s3CA

 We load the simulated data from the `data(intro)`, and get the first pair of score from s3CA, then we cluster the scores using kmeans clustreing with two clusters. We would like to check if these scores recover the target specific clusters. 

```{r s3CA}
s3CA_res <- cCCA(Xa = intro$s3CA$Xa, Ya = intro$s3CA$Ya, 
                 Xt = intro$s3CA$Xt, Yt = intro$s3CA$Yt, eta  = NULL,
                 sparse_ctst = sqrt(20/ncol(intro$s3CA$Xt)), 
                 sparse_trt = sqrt(20/ncol(intro$s3CA$Xt)), 
                 sparse_ctrl = sqrt(20/ncol(intro$s3CA$Xa)))
```


```{r s3CA-usage}
df_score <- data.frame(CCcX=intro$s3CA$Xt %*% s3CA_res$CCA_ctst$u[,1], CCcY=intro$s3CA$Yt %*% s3CA_res$CCA_ctst$v[,1], labels = intro$s3CA$labels)
clustering <- spec_clust(df_score[,1:2], k = 2, group = intro$s3CA$labels)
df_score$Clusters <- clustering$cluster
ggplot(df_score)+geom_point(aes(x = CCcX, y = CCcY, color= factor(labels), shape = factor(Clusters)), size = 2)+
  labs(title = "Contrastive", color = "True target specific clusters", shape = "Estimated clusters from contrastive")+
  theme_bw()+
  guides(
    shape = guide_legend(
      nrow = 1,
      byrow = TRUE,
      order = 1
    ),
    color = guide_legend(
      nrow = 1,
      byrow = TRUE,
      order = 2
    )
  ) +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 14),  # Adjust text size
    axis.title = element_text(size = 16),  # Adjust axis title size
    axis.text = element_text(size = 12),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.box.just = "center"
  )
cat("ARI for target specific clusters:", clustering$ari, "\n")
```