---
title: "ContrastiveSpectralClustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ContrastiveSpectralClustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 10,
  out.width = "100%"
)
```

```{r setup, include = FALSE}
suppressMessages({
  library(ConSpec)
  library(ggplot2)
  library(ggsci)
  library(ggh4x)
  library(ggpubr)
  library(tidyverse)
  library(ComplexHeatmap)
  library(RColorBrewer)
  library(Seurat)
  library(SeuratData)
  library(dplyr)
  library(viridis)
  library(cluster)
  library(patchwork)
  library(aricode)
  library(PMA)
  library(pheatmap)
  library(cowplot)
  library(compositions)
  library(gridExtra)
  library(grid)
})
select <- dplyr::select

```


# Introduction 

ConSpec is an R package for contrastive spectral clustering including contrastive PCA (cPCA) and its sparse version (scPCA), contrastive cross-covariance analysis (3CA) and its sparse version (s3CA).


In this vignette, we will use ConSpec package to reproduce the analysis in the original paper, including

1. Single dataset eta tuning example, generating Fig.2

2. Simulation for contrastive spectral clustering, generating Table 1.

3. Simulation for sparse contrastive spectral clustering, generating Table 2.

4. Simulation for 3CA and s3CA, generating Table 3

5. CRISPER screen data application of cPCA generating Figures in Section 2 and Section 7.1

6. Broiler multiomics data application of s3CA generating Figures in Section 2 and Section 7.2

# Simulation

## 1. Single dataset eta tuning example

```{r  eta-tuning-example, fig.width=5, fig.height=5, out.width="100%"}
data(sim_data_CSC)
eta_range <-  seq(0, 100, 0.2)
Xt <- scale(sim_data_CSC$Xt, scale = FALSE)
Xa <- scale(sim_data_CSC$Xa, scale = FALSE)
meta_t <- sim_data_CSC$covariates_t
eta_tuning_res <-  eta_tuning_general(Xt = Xt, Xa = Xa, eta_range = eta_range)
eta_opt <- eta_tuning_res$eta_opt
p_tuning <- eta_tuning_res$plot+scale_color_cosmic()+labs(title = expression("ARI/Silhouette vs candidate"~eta))


meta_all <-  setNames(data.frame(matrix(ncol = ncol(meta_t)+3, nrow = 0)), c(colnames(meta_t), "cPC1","cPC2","eta"))
for (eta_use in c(eta_opt, seq(from = min(eta_range), to = max(eta_range), length.out = 3))) {
  meta_tmp <- meta_t
  meta_tmp$cPC1 <-  Xt %*% eigen(cov(Xt)-eta_use*cov(Xa))$vectors[,1]
  meta_tmp$cPC2 <-  Xt %*% eigen(cov(Xt)-eta_use*cov(Xa))$vectors[,2]
  meta_tmp$eta <- paste0("eta==",eta_use)
  meta_all <- rbind(meta_all, meta_tmp)
}
meta_all$eta <- factor(meta_all$eta, levels=c("eta==0","eta==1.6","eta==50","eta==100"))
p_eta <- ggplot(meta_all)+
            geom_point(aes(x = cPC1, y = cPC2, color = interesting, shape = nuisance))+
            facet_wrap(vars(eta), labeller = label_parsed, scales= "free")+
            theme_bw()+
            scale_color_aaas()+
            scale_shape_manual(values = c(1, 15))+
            theme(
              strip.background = element_rect(fill = "lightblue", color = "black"),
              strip.text = element_text(color = "black", face = "bold", size = 14),
              text = element_text(size = 14),  # Adjust text size
              axis.title = element_text(size = 16),  # Adjust axis title size
              axis.text = element_text(size = 12),   # Adjust axis text size
              legend.position = "bottom")
print(p_eta)


## split the figures into main figure and supp figure
p_eta_main <- ggplot(meta_all[meta_all$eta %in% c("eta==0","eta==1.6"),])+
    geom_point(aes(x = cPC1, y = cPC2, color = interesting, shape = nuisance))+
    facet_wrap(vars(eta), labeller = label_parsed, scales= "free")+
    theme_bw()+
    scale_color_aaas()+
    scale_shape_manual(values = c(1, 15))+
    theme(
        strip.background = element_rect(fill = "lightblue", color = "black"),
        strip.text = element_text(color = "black", face = "bold", size = 14),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),   # Adjust axis text size
        legend.position = "bottom")

p_eta_supp <- ggplot(meta_all[!(meta_all$eta %in% c("eta==0","eta==1.6")),])+
    geom_point(aes(x = cPC1, y = cPC2, color = interesting, shape = nuisance))+
    facet_wrap(vars(eta), labeller = label_parsed, scales= "free")+
    theme_bw()+
    scale_color_aaas()+
    scale_shape_manual(values = c(1, 15))+
    theme(
        strip.background = element_rect(fill = "lightblue", color = "black"),
        strip.text = element_text(color = "black", face = "bold", size = 14),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),   # Adjust axis text size
        legend.position = "bottom")
# ggsave("../inst/SavedFigures/sim_etaTuningEg_Main.pdf",
#        plot = ggarrange(p_tuning+xlim(0,3), p_eta_main, ncol = 2, widths = c(1,1.6), 
#        labels ="AUTO"), width = 12, height = 4.5, units = "in")
# ggsave("../inst/SavedFigures/sim_etaTuningEg_Supp.pdf",
#        plot = ggarrange(p_tuning, p_eta_supp, ncol = 2, widths = c(1,1.6), 
#        labels ="AUTO"), width = 12, height = 4.5, units = "in")
```

## 2. Simulation for contrastive spectral clustering, generating Table 1 
```{r  simulation-cPCA, eval=FALSE}
set.seed(111)
mu2_candidates <- c(1, 2)
p1_candidates <- c(5, 10)
eta_range <-  seq(0.1, 50, 0.2)
Sigmat_candidates <- c("Identity", "Toep(0.4)", "Toep(0.5)")
results <- setNames(data.frame(matrix(ncol = 8, nrow = 0)),
                     c("ARI_ctst_interesting", "ARI_ctst_nuisance",
                       "ARI_trt_interesting", "ARI_trt_nuisance","eta",
                        "mu2", "p1", "Sigmat"))
for (mu2 in mu2_candidates) {
  for (p1 in p1_candidates) {
    for(Sigmat in Sigmat_candidates){
        p2 <- 10
        p <- p1+p2
        Sigma_t <- case_when(Sigmat == "Identity" ~ diag(1, p),
                             Sigmat == "Toep(0.4)" ~ toeplitz(0.4^(0:(p-1))),
                             Sigmat == "Toep(0.5)" ~ toeplitz(0.5^(0:(p-1))))
        res <- sim_CSC(n11 = 50, n12 = 25, n21 = 55, n22 = 30, m1tilde = 20, m2tilde = 30,
                       mu1 = 5, mu2 = mu2, theta1 = -5, theta2 = 6, p1 = p1, p2 = p2,
                       sigma_t = 1, sigma_a = 1,
                       Sigma_t = Sigma_t, Sigma_a = NULL,
                       eta = eta_range, repetition = 50, save_plot = FALSE)
        res$mu2 <- mu2
        res$mu <- abs(mu2-5)
        res$p1 <- p1
        res$Sigmat <- Sigmat
        results <- rbind(results, res)
      }
    }
  }
# save(results, file = "../inst/SavedData/sim_cPCA.RData")
```

```{r  simulation-cPCA-visualization, eval=TRUE}
load(file = "../inst/SavedData/sim_cPCA.RData")
plot_results <- pivot_longer(results, 1:4, names_to = "Type", values_to = "ARI")
plot_results$Analysis <- sapply(plot_results$Type, function(x){strsplit(x,"_")[[1]][2]})
plot_results$Cluster <- sapply(plot_results$Type, function(x){strsplit(x,"_")[[1]][3]})
plot_results$Analysis <- ifelse(plot_results$Analysis=="ctst","Contrastive","Target-only")
plot_results$Cluster <- ifelse(plot_results$Cluster=="interesting","Interesting","Nuisance")
plot_results$eta <- paste("eta==", plot_results$eta, sep = "")
plot_results$mu <- paste("mu==", plot_results$mu, sep = "")
plot_results$p1 <- paste("p[1]==", plot_results$p1, sep = "")
plot_results <- plot_results %>% group_by(mu, p1, Sigmat, Analysis, Cluster) %>%
  summarise(Average_ARI = mean(ARI), SE_ARI = sd(ARI), covarage_high = mean(ARI >= 0.9), covarage_low = mean(ARI <= 0.1))
print(knitr::kable(plot_results, format = "markdown", digits = 3))
p_cPCA <- ggplot(plot_results, aes(x = Cluster,  y = Average_ARI))+
          geom_point(aes(color = Analysis), stat="identity", position ="dodge")+
          geom_errorbar(aes(ymin = Average_ARI - SE_ARI, ymax = Average_ARI + SE_ARI), linewidth = 0.5) +
          facet_nested(Sigmat ~ mu + p1, labeller = label_parsed)+
          labs(x = "", y="Average ARI")+
          theme_bw()+
          theme(
            panel.grid = element_blank(),
            strip.background = element_rect(fill = "lightblue", color = "black"),
            strip.text = element_text(color = "black", face = "bold", size = 14),
            text = element_text(size = 14),  # Adjust text size
            axis.title = element_text(size = 16),  # Adjust axis title size
            axis.text = element_text(size = 12),   # Adjust axis text size
            legend.position = "bottom"
         )+
         scale_color_aaas()
print(p_cPCA)
# ggsave("../inst/SavedFigures/sim_cPCA.pdf", p_cPCA, width = 6, height = 9, units = "in")
```

## 3. Simulation for sparse contrastive spectral clustering, generating Table 2

```{r simulation-scPCA, eval = FALSE}
set.seed(111)
mu2_candidates <- c(1, 2)
p3_candidates <- c(100, 300)
eta_range <-  seq(0.1, 10, 0.5)
Sigmat_candidates <- c("Identity", "Toep(0.4)", "Toep(0.5)")
results <- setNames(data.frame(matrix(ncol = 8, nrow = 0)),
                     c("ARI_ctst_interesting", "ARI_ctst_nuisance",
                     "ARI_trt_interesting", "ARI_trt_nuisance","eta",
                      "mu2", "p3", "Sigmat"))
for (mu2 in mu2_candidates) {
  for (p3 in p3_candidates) {
    for(Sigmat in Sigmat_candidates){
      p1 <- 5
      p2 <- 10
      p <- p1+p2+p3
      Sigma_t <- case_when(Sigmat == "Identity" ~ diag(1, p),
                           Sigmat == "Toep(0.4)" ~ toeplitz(0.4^(0:(p-1))),
                           Sigmat == "Toep(0.5)" ~ toeplitz(0.5^(0:(p-1))))
      res <- sim_sCSC(n11 = 50, n12 = 25, n21 = 55, n22 = 30, m1tilde = 20, m2tilde = 30,
                      mu1 = 5, mu2 = mu2, theta1 = -5, theta2 = 6, p1 = p1, p2 = p2, p3 = p3,
                      sigma_t = 1, sigma_a = 1, Sigma_t = Sigma_t, Sigma_a = NULL,
                      eta = eta_range, repetition = 50)
       res$mu2 <- mu2
       res$mu <- abs(mu2-5)
       res$p3 <- p3
       res$Sigmat <- Sigmat
       results <- rbind(results, res)
     }
   }
}
# save(results, file = "../inst/SavedData/sim_scPCA.RData")
```

```{r  simulation-scPCA-visualization, eval=TRUE}
load(file = "../inst/SavedData/sim_scPCA.RData")
plot_results <- pivot_longer(results, 1:4, names_to = "Type", values_to = "ARI")
plot_results$Analysis <- sapply(plot_results$Type, function(x){strsplit(x,"_")[[1]][2]})
plot_results$Cluster <- sapply(plot_results$Type, function(x){strsplit(x,"_")[[1]][3]})
plot_results$Analysis <- ifelse(plot_results$Analysis=="ctst","Contrastive","Target-only")
plot_results$Cluster <- ifelse(plot_results$Cluster=="interesting","Interesting","Nuisance")
plot_results$eta <- paste("eta==", plot_results$eta, sep = "")
plot_results$mu <- paste("mu==", plot_results$mu, sep = "")
plot_results$p3 <- paste("p[3]==", plot_results$p3, sep = "")
plot_results <- plot_results %>% group_by(mu, p3, Sigmat, Analysis, Cluster) %>%
  summarise(Average_ARI = mean(ARI), SE_ARI = sd(ARI), covarage_high = mean(ARI >= 0.9), covarage_low = mean(ARI <= 0.1))
print(knitr::kable(plot_results, format = "markdown", digits = 3))
p_scPCA <- ggplot(plot_results, aes(x = Cluster,  y = Average_ARI))+
          geom_point(aes(color = Analysis), stat="identity", position ="dodge")+
          geom_errorbar(aes(ymin = Average_ARI - SE_ARI, ymax = Average_ARI + SE_ARI), linewidth = 0.5) +
          facet_nested(Sigmat ~ mu + p3, labeller = label_parsed)+
          labs(x = "", y="Average ARI")+
          theme_bw()+
          theme(
            panel.grid = element_blank(),
            strip.background = element_rect(fill = "lightblue", color = "black"),
            strip.text = element_text(color = "black", face = "bold", size = 14),
            text = element_text(size = 14),  # Adjust text size
            axis.title = element_text(size = 16),  # Adjust axis title size
            axis.text = element_text(size = 12),   # Adjust axis text size
            legend.position = "bottom"
         )+
         scale_color_aaas()
print(p_scPCA)
# ggsave("../inst/SavedFigures/sim_scPCA.pdf", p_scPCA, width = 6, height = 9, units = "in")
```


## 4. Simulation for 3CA and s3CA, generating Table 3

```{r simulation-3CA-and-s3CA, eval = FALSE}
set.seed(2024)
## Passing a vector to `eta` selects the contrast parameter with eta_tuning_bimodal();
## the spectral-scale ratio is still recorded in the `Eta_Ratio` column.
eta_range <- seq(0.1, 6, 0.1)
all_results <- list()
scenarios_dense <- expand.grid(p_val = c(40, 80), mu_val = c(1.5, 3))

for(i in 1:nrow(scenarios_dense)) {
  p_val <- scenarios_dense$p_val[i]
  mu_val <- scenarios_dense$mu_val[i]
  
  cat(sprintf("\nRunning 3CA (p = q = %d, mu = %.1f)...\n", p_val, mu_val))
  
  dense_res <- sim_3CA(nt = 200, na = 200, p = p_val, q = p_val, mu = mu_val,
                       theta = 400, eta_true = 2.5, repetition = 50,
                       eta = eta_range)
  
  dense_res$Method <- "3CA"
  dense_res$Dimension_p <- p_val
  dense_res$mu_or_s <- mu_val 
  
  dense_res$TPR_Target <- NA; dense_res$FPR_Target <- NA
  dense_res$TPR_Contrastive <- NA; dense_res$FPR_Contrastive <- NA
  
  all_results[[paste0("dense_p", p_val, "_mu", mu_val)]] <- dense_res
}

scenarios_sparse <- expand.grid(p_val = c(300, 500), s_val = c(10, 20))

for(i in 1:nrow(scenarios_sparse)) {
  p_val <- scenarios_sparse$p_val[i]
  s_val <- scenarios_sparse$s_val[i]
  
  cat(sprintf("Running s3CA (p = q = %d, number of non-zero elements s = %d)...\n", p_val, s_val))
  
  
    sparse_res <- sim_s3CA(nt = 200, na = 200, p = p_val, q = p_val, s = s_val,
                         mu = 3, theta = 400, eta_true = 2.5, repetition = 50,
                         eta = eta_range)
  
  sparse_res$Method <- "s3CA"
  sparse_res$Dimension_p <- p_val
  sparse_res$mu_or_s <- s_val
  
  all_results[[paste0("sparse_p", p_val, "_s", s_val)]] <- sparse_res
}

final_data <- do.call(rbind, all_results)
save(final_data, file = "../inst/SavedData/sim_3CA_s3CA.RData")
```

```{r simulation-3CA-and-s3CA-visualization, eval = TRUE}
load(file = "../inst/SavedData/sim_3CA_s3CA.RData")
summary_table <- final_data %>%
  group_by(Method, Dimension_p, mu_or_s) %>%
  summarize(
    Mean_Eta = mean(Eta_Est),
    SD_Eta = sd(Eta_Est),
    `Eta_Hat_SD` = sprintf("%.3f (%.3f)", Mean_Eta, SD_Eta),
    `Eta_Ratio_SD` = sprintf("%.3f (%.3f)", mean(Eta_Ratio), sd(Eta_Ratio)),
    ARI_TargetOnly = mean(ARI_Target),
    ARI_Contrastive = mean(ARI_Contrastive),
    TPR_TargetOnly = mean(TPR_Target, na.rm = TRUE),
    FPR_TargetOnly = mean(FPR_Target, na.rm = TRUE),
    TPR_Contrastive = mean(TPR_Contrastive, na.rm = TRUE),
    FPR_Contrastive = mean(FPR_Contrastive, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::select(-Mean_Eta, -SD_Eta) %>%
  arrange(desc(Method), Dimension_p, mu_or_s)

print(knitr::kable(summary_table, digits = 3, format = "markdown"))
```

# Real Data Applications

## 1. CRISPER screen data application of cPCA 


### A: Dataset Intro

Papalexi E, Mimitou E P, Butler A W, et al. Characterizing the molecular regulation of inhibitory immune checkpoints with multimodal single-cell screens[J]. Nature genetics, 2021, 53(3): 322-331.

This dataset comes from an ECCITE-seq pooled CRISPR-Cas9 screen in IFN-γ-stimulated THP-1 cells. For each single cell, it jointly measures gene expression, surface protein abundance including PD-L1, and the assigned gRNA perturbation. The screen targets genes involved in IFN-γ signaling and immune checkpoint regulation, with non-targeting controls, making it useful for studying how genetic perturbations affect transcriptomic and PD-L1 protein responses while accounting for confounders such as cell cycle, batch effects, and cellular stress.

### B: Methods

#### B-1. Data Preprocessing and Feature Selection

Raw counts were normalized using standard log-normalization for RNA and Centered Log-Ratio (CLR) normalization for ADT. Variable features (n=1,000) were identified, and data were scaled using the Seurat package to zero mean and unit variance.

#### B-2. cPCA

Standard PCA revealed that the primary source of variation in the dataset was the cell cycle phase, which confounded the biological signals associated with CRISPR perturbations. To correct for this, we applied Contrastive PCA (cPCA). cPCA identifies a low-dimensional subspace that maximizes the variance in a target dataset (cells with specific gene knockouts) while minimizing the variance in a background dataset (Non-Targeting control cells).

We defined the **Target** group ($X_t$) as cells carrying gRNAs for IFN pathway components (*IFNGR1*, *IFNGR2*, *JAK2*, *STAT1*, *IRF1*) and the **Ancillary** group ($X_a$) as cells carrying Non-Targeting (NT) gRNAs. The cPCA solution was obtained by performing eigenvalue decomposition on the contrastive covariance matrix:

$$ \Sigma_{cPCA} = \text{Cor}(X_t) - \eta \times \text{Cor}(X_a) $$

A contrastive hyperparameter of $\eta = 5$ was selected to effectively suppress cell-cycle-driven variance. The resulting contrastive Principal Components (cPCs) were used for visualization and k-means clustering ($k=3$).

#### B-3. Validation of Clusters

To validate the biological relevance of the cPCA clusters, we employed two orthogonal methods:

1.  **Protein Validation:** We mapped the expression of PD-L1  surface protein onto the cPCA latent space. Since PD-L1 expression is upregulated by the IFN-$\gamma$ pathway, true genetic knockouts of pathway components should exhibit reduced PD-L1 levels compared to wild-type cells.

2.  **Perturbation Signature Classification (Mixscape):** We utilized the Mixscape algorithm (Papalexi et al., 2021) to classify cells into "Perturbed" (KO) or "Non-Perturbed/Escape" (NP) states. This classification was projected onto the cPCA space to assess whether cPCA correctly resolved phenotypic heterogeneity.

### C. Results and Findings

#### C-1. cPCA Resolves Pathway Architecture and Removes Confounders
While standard PCA was dominated by cell cycle effects (G1/S/G2M separation), cPCA successfully removed this background variation, revealing a structured "L-shaped" topology in the latent space. The analysis identified three distinct states:

1.  **Cluster 1 (Upstream Blockade):** Composed of cells perturbed for *IFNGR1*, *IFNGR2*, *JAK2*, and *STAT1*. These genes represent the receptor and immediate signaling complex of the IFN-$\gamma$ pathway.

2.  **Cluster 2 (Downstream Blockade):** Composed of cells perturbed for *IRF1*, a transcription factor downstream of the JAK-STAT complex, indicating a distinct transcriptomic response compared to upstream receptor blockade.

3.  **Cluster 3 (Central/Wild-Type State):** A central cluster overlapping with the origin.

#### C-2. Identification of Escape Cells and Phenotypic Heterogeneity

A significant proportion of cells carrying targeting gRNAs clustered in the central region, indistinguishable from the Non-Targeting baseline. We hypothesized these were "escape" cells that harbored the gRNA but failed to undergo functional genetic knockout.

*   **Mixscape Confirmation:** Projection of Mixscape classifications confirmed that the "independent" clusters (Upstream and Downstream) were populated by cells classified as "Perturbed (KO)" (red points), while the central overlapping region was populated by cells classified as "Escape (NP)" (grey points).

*   **Protein Validation:** Orthogonal ADT data demonstrated that cells in the Upstream and Downstream clusters exhibited significantly reduced PD-L1 protein levels, whereas cells in the central overlap region maintained high PD-L1 levels comparable to wild-type controls.


```{r CRISPER-PCA, eval = TRUE, fig.width=10, fig.height=5, out.width="100%"}
# ==============================================================================
# Module 1: Setup and Data Loading
# ==============================================================================
# Load Data
data("thp1.eccite")
eccite <- UpdateSeuratObject(thp1.eccite)

# ==============================================================================
# Module 2: Preprocessing (RNA & ADT)
# Standard Seurat workflow to normalize and scale data
# ==============================================================================
# 1. RNA Processing
DefaultAssay(eccite) <- "RNA"
eccite <- NormalizeData(eccite) %>% 
  FindVariableFeatures(nfeatures = 1000) %>% 
  ScaleData() %>% 
  RunPCA(verbose = FALSE) %>% 
  RunUMAP(dims = 1:40, verbose = FALSE)

# 2. ADT (Protein) Processing - CLR Normalization
DefaultAssay(eccite) <- "ADT"
eccite <- NormalizeData(eccite, normalization.method = "CLR", margin = 2) %>% 
  ScaleData()

# 3. Extract Scaled Matrices for cPCA
# Transpose to N_cells x N_genes format
rna_matrix_final <- t(GetAssayData(eccite, assay = "RNA", layer = "scale.data"))
adt_matrix_final <- t(GetAssayData(eccite, assay = "ADT", layer = "scale.data"))

# Ensure matrices are dense and aligned
rna_matrix_final <- as.matrix(rna_matrix_final)
adt_matrix_final <- as.matrix(adt_matrix_final)
meta_data <- eccite@meta.data

# Sanity Check: Alignment
stopifnot(all(rownames(rna_matrix_final) == rownames(adt_matrix_final)))
stopifnot(all(rownames(meta_data) == rownames(rna_matrix_final)))

# ==============================================================================
# Module 3: Contrastive PCA (cPCA) Setup & Execution
# Goal: Isolate perturbation signals by removing background (NT) variation
# ==============================================================================

# 1. Define Subgroups
# Target Group (Xt): Cells with specific gene knockouts
# Background Group (Xa): Non-Targeting (NT) control cells
target_genes <- c("IFNGR1", "IFNGR2", "JAK2", "STAT1", "IRF1")
background_gene <- "NT"

# Filter Metadata
sub_meta <- meta_data %>% 
  filter(gene %in% c(target_genes, background_gene)) %>%
  dplyr::select(Treatment_Gene = gene, Treatment_Guide = guide_ID, Phase, Replicate = replicate)

# Filter Matrix
sub_matrix <- rna_matrix_final[rownames(sub_meta), ]

# Split into Target (Xt) and Background (Xa)
Xt_raw <- sub_matrix[sub_meta$Treatment_Gene != "NT", ]
meta_t <- sub_meta[sub_meta$Treatment_Gene != "NT",]

Xa_raw <- sub_matrix[sub_meta$Treatment_Gene == "NT", ]
meta_a <- sub_meta[sub_meta$Treatment_Gene == "NT",]

# 2. Feature Selection
# Remove genes with zero variance in either group to prevent division errors
keep_genes <- apply(Xt_raw, 2, var) > 0 & apply(Xa_raw, 2, var) > 0
Xt <- scale(Xt_raw[, keep_genes])
Xa <- scale(Xa_raw[, keep_genes])

print(paste("Features retained:", sum(keep_genes)))

# 3. Run PCA on each group
pca_t <- prcomp(Xt, center = TRUE, scale. = TRUE)
pca_a <- prcomp(Xa, center = TRUE, scale. = TRUE)

plot_df_t <- cbind(as.data.frame(pca_t$x), meta_t)
plot_df_a <- cbind(as.data.frame(pca_a$x), meta_a)
p_t <- ggplot(plot_df_t, aes(x = PC1, y = PC2, color = Phase, shape = Treatment_Gene)) +
  geom_point(alpha = 0.6, size = 2) +
  theme_bw() +
  labs(title = "Target-only", shape = "Perturbed genes")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "bottom")

p_a <- ggplot(plot_df_a, aes(x = PC1, y = PC2, color = Phase)) +
  geom_point(alpha = 0.6, size = 2, shape = 6) +
  theme_bw() +
    labs(title = "Ancillary-only")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "bottom")

print(ggarrange(p_t,p_a, common.legend = TRUE,legend = "bottom", labels = "AUTO", align = "h"))
# p_motivation_1 <- ggplot(plot_df_t, aes(x = PC1, y = PC2, color = Phase, shape = Treatment_Gene)) +
#   geom_point(alpha = 0.6, size = 2) +
#   theme_bw() +
#   labs(shape = "Perturbed genes")+
#   theme(panel.grid = element_blank(),
#         text = element_text(size = 14),  # Adjust text size
#         axis.title = element_text(size = 16),  # Adjust axis title size
#         axis.text = element_text(size = 12),
#         legend.position = "bottom",
#         legend.box      = "vertical")   # stack color legend over shape legend)+
#    guides(
#     color = guide_legend(nrow = 1, byrow = TRUE, order = 1),
#     shape = guide_legend(nrow = 1, byrow = TRUE, order = 2)
#   )

# ggsave("../inst/SavedFigures/ECCITE_score_ta.pdf", 
#        plot = ggarrange(p_t,p_a, common.legend = TRUE,legend = "bottom", labels = "AUTO", align = "h"), 
#        width = 10, height = 5)



```


```{r CRISPER-eta-tuning, eval = TRUE, fig.width=5, fig.height=4, out.width="100%"}
# Xt and Xa are already centered and scaled above, so cov() inside eta_tuning_general()
# operates on the correlation matrices, matching cor(Xt) below.
# This application looks for three contrastive clusters (Upstream, Downstream, Escape),
# so ckm_cluster = 3 and the default clust_comps = "auto" clusters on ckm_cluster - 1 = 2
# cPC scores.
set.seed(111)
eta_tuning_crispr <- eta_tuning_general(
  Xt = Xt, Xa = Xa,
  eta_range = seq(0.5, 12, 0.25),
  plot = TRUE,
  sparse_trt = NULL, sparse_ctst = NULL,
  km_cluster = 2, ckm_cluster = 3
)
eta_opt <- eta_tuning_crispr$eta_opt
cat("Selected eta:", eta_opt, "\n")
```


```{r CRISPER-cPCA, eval = TRUE, fig.width=5, fig.height=5, out.width="100%"}
# 4. Run cPCA Algorithm
# Formula: Eigen decomposition of (Cor(Xt) - eta * Cor(Xa)), with eta from the chunk above
contrastive_cov <- cor(Xt) - eta_opt * cor(Xa)
cpca_res <- eigen(contrastive_cov)

# 4. Project Data onto Contrastive Principal Components (cPCs)
cpca_coords <- as.data.frame(Xt %*% cpca_res$vectors[, 1:2])
colnames(cpca_coords) <- c("cPC1", "cPC2")

# Merge with Metadata
cpca_df <- cbind(cpca_coords, sub_meta[rownames(Xt), ])

# ==============================================================================
# Module 4: Clustering and Functional Annotation
# ==============================================================================
set.seed(123) # Reproducibility for K-means
cpca_df$Clusters <- factor(kmeans(cpca_df[, c("cPC1","cPC2")], centers = 3)$cluster)

# Annotate Functional Groups for Validation
cpca_df <- cpca_df %>%
  mutate(Functional = case_when(
    Treatment_Gene == "IRF1" ~ "Downstream",
    Treatment_Gene %in% c("IFNGR1", "IFNGR2", "JAK2", "STAT1") ~ "Upstream",
    TRUE ~ "Other" ),
    Functional = factor(Functional,
                        levels = c("Upstream", "Downstream", "Other"))
    )

# Visualization: The cPCA Space
p_c <- ggplot(cpca_df, aes(x = cPC1, y = cPC2, colour = Functional, shape = Clusters)) +
  geom_point(alpha = 0.6, size = 2) +
  theme_bw() +
  scale_color_aaas()+
  labs(shape = "Clusters", colour = "Functional")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "bottom")

print(p_c)

ggsave("../inst/SavedFigures/ECCITE_score_tac.pdf", 
       plot = ggarrange(p_t+theme(legend.box = "vertical"),
                        p_a+theme(legend.box = "vertical"),
                        p_c+theme(legend.box = "vertical")+labs(title = "Contrastive"), 
                        labels = "AUTO", align = "h", nrow = 1), 
       width = 19, height = 8)

# ==============================================================================
# Module 5: Validation 1 - Protein Data (ADT)
# Hypothesis: True KO cells should have low PD-L1 protein expression
# ==============================================================================
# Map PD-L1 expression to cPCA coordinates
cpca_df$PDL1_Protein <- adt_matrix_final[rownames(cpca_df), "PDL1"]

p_val_protein <- ggplot(cpca_df, aes(x = cPC1, y = cPC2, color = PDL1_Protein)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_viridis(option = "C", name = "PD-L1 (CLR Scaled)") + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "bottom")

print(p_val_protein)

# ==============================================================================
# Module 6: Validation 2 - Mixscape Analysis (Perturbation Signature)
# Hypothesis: Center overlap region contains 'Escape' cells (Wild-type phenotype)
# ==============================================================================
# 1. Calculate Perturbation Signature
eccite <- CalcPerturbSig(
  object = eccite, 
  assay = "RNA", 
  slot = "data", 
  gd.class ="gene", 
  nt.cell.class = "NT", 
  reduction = "pca", 
  ndims = 40, 
  num.neighbors = 20, 
  split.by = "replicate",
  new.assay.name = "PRTB",
  verbose = FALSE
)

# 2. Run Mixscape Classification
DefaultAssay(eccite) <- 'PRTB'
VariableFeatures(object = eccite) <- VariableFeatures(object = eccite[["RNA"]])
eccite <- ScaleData(eccite, do.scale = F, do.center = T) %>%
  RunPCA(reduction.name = 'prtbpca', reduction.key = "prtbpca" ,verbose = FALSE) %>%
  RunUMAP(dims = 1:40, reduction = 'prtbpca', reduction.name = 'prtbumap', reduction.key = 'prtbumap', verbose = FALSE)

eccite <- RunMixscape(
  object = eccite, 
  assay = "PRTB", 
  slot = "scale.data", 
  labels = "gene", 
  nt.class.name = "NT", 
  min.de.genes = 5, 
  iter.num = 10, 
  de.assay = "RNA", 
  prtb.type = "KO",
  verbose = FALSE
)

# 3. Map Mixscape Labels to cPCA
mixscape_labels <- eccite$mixscape_class[rownames(cpca_df)]
cpca_df$Mixscape_Class <- mixscape_labels

# Simplify Labels for Plotting
cpca_df <- cpca_df %>%
  mutate(Mixscape_Status = case_when(
    grepl("KO", Mixscape_Class) ~ "Perturbed",
    grepl("NP", Mixscape_Class) ~ "Escaped",
    TRUE ~ "Other"
  ))

p_val_mixscape <- ggplot(cpca_df, aes(x = cPC1, y = cPC2, color = Mixscape_Status)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = c("Perturbed" = "#d62728", 
                                "Escaped" = "grey80")) +  
  theme_bw() +
  labs(color = "Mixscape status")+
  theme(panel.grid = element_blank(),
        text = element_text(size = 14),  # Adjust text size
        axis.title = element_text(size = 16),  # Adjust axis title size
        axis.text = element_text(size = 12),
        legend.position = "bottom") 

print(p_val_mixscape)


# ggsave("../inst/SavedFigures/ECCITE_val.pdf", 
#        plot = ggarrange(p_val_protein, p_val_mixscape,  
#                         labels = "AUTO", align = "h", nrow = 1), 
#        width = 15, height = 8)

# ggsave("../inst/SavedFigures/ECCITE_c_val.pdf", 
#        plot = ggarrange(p_c+theme(legend.box = "vertical"),
#         p_val_protein, p_val_mixscape,  
#                         labels = "AUTO", align = "h", nrow = 1), 
#        width = 15, height = 8)
```


### D. Sensitivity analysis of tuning parameter

To show that the cPCA results are robust to the choice of the contrastive hyperparameter ($\eta$), we performed a sensitivity analysis by varying it across a range of values (1, 2, 3, 4, 5, 6, 8, 10). For each candidate value, we re-ran the cPCA algorithm and generated two plots, one is the cPCA latent space colored by functional annotation and shaped by k-means clusters, and the other is the PD-L1 protein expression mapped onto the cPCA space. This allowed us to assess whether the biological signal of interest (IFN pathway perturbations) remains consistently resolved across different $\eta$ values and whether the expected validation (reduced PD-L1 in true KO clusters) holds consistently.


```{r CRISPER-eta-sensitivity, eval = TRUE}
set.seed(123)
# 1. Define Candidates for Sensitivity Analysis
eta_candidates <- c(1, 2, 3, 4, 5, 6, 8, 10) 

# Lists to store plots
plots_row1_cpca <- list()    # For Functional Clusters
plots_row2_protein <- list() # For PD-L1 Protein

# 2. Loop through each Eta value
for (eta_val in eta_candidates) {
  
  # --- A. Run cPCA Algorithm ---
  contrastive_cov <- cor(Xt) - eta_val * cor(Xa)
  cpca_res <- eigen(contrastive_cov)
  
  # --- B. Project Data ---
  cpca_coords <- as.data.frame(Xt %*% cpca_res$vectors[, 1:2])
  colnames(cpca_coords) <- c("cPC1", "cPC2")
  
  # --- C. Prepare Dataframe ---
  # Note: We match rows by name to ensure alignment
  curr_df <- cbind(cpca_coords, sub_meta[rownames(Xt), ])
  
  # Add Functional Annotation
  curr_df$Clusters <- factor(kmeans(curr_df[, c("cPC1","cPC2")], centers = 3)$cluster)

  curr_df <- curr_df %>%
    mutate(Functional = case_when(
      Treatment_Gene == "IRF1" ~ "Downstream",
      Treatment_Gene %in% c("IFNGR1", "IFNGR2", "JAK2", "STAT1") ~ "Upstream",
      TRUE ~ "Other"),
      Functional = factor(Functional, levels = c("Upstream", "Downstream", "Other")))
  
  # Add Protein Data
  # Ensure we grab the correct rows
  curr_df$PDL1_Protein <- adt_matrix_final[rownames(curr_df), "PDL1"]
  
  # --- D. Generate Plot 1: Functional Clusters (Row 1) ---
  p_cpca <- ggplot(curr_df, aes(x = cPC1, y = cPC2, colour = Functional, shape = Clusters)) +
    geom_point(alpha = 0.6, size = 2) +
    theme_bw() +
    labs(title = bquote(eta == .(eta_val)),
         subtitle = NULL) +
    theme(legend.position = "none", # Hide legend for grid clarity
          axis.title = element_blank(),
          plot.title = element_text(size=11, face="bold"))+scale_color_aaas()
  
  plots_row1_cpca[[as.character(eta_val)]] <- p_cpca
  
  # --- E. Generate Plot 2: Protein Validation (Row 2) ---
  p_protein <- ggplot(curr_df, aes(x = cPC1, y = cPC2, color = PDL1_Protein)) +
    geom_point(alpha = 0.8, size = 2) +
    scale_color_viridis(option = "C") + 
    theme_bw() +
    labs(title = bquote(eta == .(eta_val)),
         subtitle = NULL) +
    theme(legend.position = "none", # Hide legend for grid clarity
          axis.title = element_blank(),
          plot.title = element_text(size=11, face="bold"))
  
  plots_row2_protein[[as.character(eta_val)]] <- p_protein
}

# ==============================================================================
# Arrange 8 Figures (4 Columns x 4 Rows)
# ==============================================================================
# Use patchwork to arrange
final_grid <- wrap_plots(c(plots_row1_cpca, plots_row2_protein), ncol = 4, nrow = 4) 

print(final_grid)

# ggsave("../inst/SavedFigures/ECCITE_eta_sensitivity.pdf", 
#        plot = final_grid, 
#        width = 12, height = 12)
```

### E. Baseline Methods for Comparison

To evaluate the efficacy of Contrastive PCA (cPCA), we compared it against two standard nuisance regression approaches. These methods represent alternative strategies for handling dominant background variation (such as the cell cycle) without using the specific contrastive optimization function of cPCA.

#### E-1. Target-Only Nuisance Regression
This method represents a standard "unsupervised" approach often used when control data is unavailable or disregarded during the feature extraction phase. It operates on the assumption that the dominant sources of variation in the target dataset ($X_t$) are technical confounders, while the biological signal of interest is contained in lower-variance components.

**Implementation:**

1.  Perform standard PCA on the target data matrix $X_t$.

2.  Identify the top $k$ principal components (e.g., $k=5$) under the assumption that they capture the dominant cell cycle effect

3.  Reconstruct the data matrix using only these top $k$ components to create a "noise approximation."

4.  Subtract this noise approximation from the original data to obtain residual data ($X_{resid}$).

5.  Perform PCA on $X_{resid}$ to visualize the remaining variation.

**Potential Drawbacks:**

This method relies entirely on variance magnitude. If the biological perturbation (e.g., *IFNGR1* knockout) induces a strong transcriptomic change, that signal will likely be captured in the top principal components. Regressing out the top $k$ PCs would inadvertently remove the primary biological signal of interest alongside the confounders. In other words, it's difficult to determine a correct $k$ that removes nuisance variation without also removing the interesting biological signal.

#### E-2. Ancillary-Based Nuisance Regression 

This method is a "supervised" extension of nuisance regression. Instead of guessing which components in the target data are noise, it utilizes the ancillary (Non-Targeting) data ($X_a$) to explicitly learn the structure of the background variation.

**Implementation:**

1.  Perform standard PCA on the ancillary data matrix $X_a$ to obtain the eigenvectors (loadings) representing the background structure (e.g., the cell cycle axes).

2.  Select the top $k$ eigenvectors ($V_a$) that explain the majority of the variance in $X_a$.

3.  Project the target data $X_t$ onto these ancillary eigenvectors to calculate the "background score" for each target cell.

4.  Reconstruct the background component within the target data and subtract it to obtain $X_{resid}$.

5.  Perform PCA on $X_{resid}$.

**Potential Drawbacks:**

Same to target-based nuisance regression, the success of this method depends heavily on the choice of $k$ (the number of PCs to remove). Removing too few leaves residual cell cycle noise; removing too many may subtract dimensions that are shared between the cell cycle and the treatment response, potentially distorting the biological signal.

In addition, unlike cPCA, which uses a continuous hyperparameter ($\eta$) to weigh the trade-off between target and background variance, this method performs orthogonal projection. It forces the resulting signal to be strictly orthogonal to the control's variation, which may be too restrictive if the biological signal is not perfectly orthogonal to the cell cycle.


```{r CRISPER-comparison-to-simple-alternatives, eval = TRUE, fig.width=10, fig.height=7, out.width="100%"}
set.seed(123)
# Define the K values to iterate over
k_values <- c(1, 2, 5, 10)

# Lists to store plots
plots_m1 <- list() # Target-Only
plots_m2 <- list() # Ancillary-Based

# Pre-calculate PCAs needed for the loop to save time
pca_t_full <- prcomp(Xt, center = FALSE, scale. = FALSE)
pca_a_full <- prcomp(Xa, center = FALSE, scale. = FALSE)

# Loop through each k value
for (k in k_values) {
  
  # ==========================================================================
  # Method 1: Target-Only Nuisance Regression
  # ==========================================================================
  # 1. Identify Noise (Top k PCs of Target)
  loadings_noise_t <- pca_t_full$rotation[, 1:k, drop = FALSE] # drop=FALSE prevents vector conversion if k=1
  
  # 2. Reconstruct Noise & Subtract
  # Noise = (Xt * Loadings) * t(Loadings)
  scores_noise_t <- Xt %*% loadings_noise_t
  noise_matrix_t <- scores_noise_t %*% t(loadings_noise_t)
  Xt_resid_1 <- Xt - noise_matrix_t
  
  # 3. Final PCA on Residuals
  pca_m1 <- prcomp(Xt_resid_1, center = FALSE, scale. = FALSE)
  
  # 4. Prepare Data
  df_m1 <- data.frame(
    PC1 = pca_m1$x[,1],
    PC2 = pca_m1$x[,2],
    Treatment_Gene = meta_t$Treatment_Gene,
    Phase = meta_t$Phase
  ) %>%
    mutate(Functional = case_when(
      Treatment_Gene == "IRF1" ~ "Downstream",
      Treatment_Gene %in% c("IFNGR1", "IFNGR2", "JAK2", "STAT1") ~ "Upstream",
      TRUE ~ "Other"),
      Functional = factor(Functional, levels = c("Upstream", "Downstream", "Other"))
    )
  
  df_m1$Clusters <- factor(kmeans(df_m1[, c("PC1","PC2")], centers = 3)$cluster)

  # 5. Plot
  p1 <- ggplot(df_m1, aes(x = PC1, y = PC2, color = Functional, shape = Phase)) +
    geom_point(alpha = 0.6, size = 2) +
    theme_bw() +
    labs(title = bquote("Target-based (" * kappa == .(k) * ")"),
         subtitle = NULL) +
    theme(legend.position = "none",
          axis.title = element_blank(), # Clean up axes for grid
          plot.title = element_text(size=10))+
    scale_color_aaas()+
    scale_shape_manual(
    values = c(
      "G1" = 1,  
      "G2M" = 2,  
      "S" = 0   
    ))
  
  plots_m1[[as.character(k)]] <- p1
  
  # ==========================================================================
  # Method 2: Ancillary-Based Nuisance Regression
  # ==========================================================================
  # 1. Extract Noise Vectors (Top k PCs of Ancillary/NT)
  vectors_noise_a <- pca_a_full$rotation[, 1:k, drop = FALSE]
  
  # 2. Project Target onto Noise Vectors & Subtract
  projection_scores <- Xt %*% vectors_noise_a
  noise_in_Xt <- projection_scores %*% t(vectors_noise_a)
  Xt_resid_2 <- Xt - noise_in_Xt
  
  # 3. Final PCA on Residuals
  pca_m2 <- prcomp(Xt_resid_2, center = FALSE, scale. = FALSE)
  
  # 4. Prepare Data
  df_m2 <- data.frame(
    PC1 = pca_m2$x[,1],
    PC2 = pca_m2$x[,2],
    Treatment_Gene = meta_t$Treatment_Gene,
    Phase = meta_t$Phase
  ) %>%
    mutate(Functional = case_when(
      Treatment_Gene == "IRF1" ~ "Downstream",
      Treatment_Gene %in% c("IFNGR1", "IFNGR2", "JAK2", "STAT1") ~ "Upstream",
      TRUE ~ "Other"),
      Functional = factor(Functional, levels = c("Upstream", "Downstream", "Other"))
    )
  
  df_m2$Clusters <- factor(kmeans(df_m2[, c("PC1","PC2")], centers = 3)$cluster)
  
  # 5. Plot
  p2 <- ggplot(df_m2, aes(x = PC1, y = PC2, color = Functional, shape = Phase)) +
    geom_point(alpha = 0.6, size = 2) +
    theme_bw() +
    labs(title = bquote("Ancillary-based (" * kappa == .(k) * ")"),
         subtitle = NULL) +
    theme(legend.position = "none", # Hide legend for grid, add back later if needed
          axis.title = element_blank(),
          plot.title = element_text(size=10))+
    scale_color_aaas()+
    scale_shape_manual(
    values = c(
      "G1" = 1,  
      "G2M" = 2,  
      "S" = 0   
    ))
  plots_m2[[as.character(k)]] <- p2
}

# ==============================================================================
# Arrange Figures (4 Columns x 2 Rows)
# Row 1: Target-Only (k=1, 2, 5, 10)
# Row 2: Ancillary-Based (k=1, 2, 5, 10)
# ==============================================================================

# Combine all plots into a single list in the correct order
all_plots_comp <- c(plots_m1, plots_m2)

# Use patchwork to arrange
# layout: 2 rows, 4 columns
final_grid_comp <- wrap_plots(all_plots_comp, ncol = 4, nrow = 2) 

print(final_grid_comp)
# ggsave("../inst/SavedFigures/ECCITE_comparison.pdf", 
#        plot = final_grid_comp, 
#        width = 12, height = 8)
```



## 2. Broiler chicken data application of s3CA  

### A. Background & Study Design

We used a **published longitudinal chicken gut microbiome dataset** to study antibiotic-associated changes during *Campylobacter jejuni* infection. The original experiment examined whether **enrofloxacin treatment** alters the gut microbial and metabolic environment in ways that may favor the persistence of resistant *C. jejuni*.

- **Biological context:**  
  Broiler chickens infected with *C. jejuni*.

- **Study objective:**  
  To evaluate how **enrofloxacin treatment** affects gut microbial and metabolic dynamics during infection.

- **Experimental design:**  
  - **92 newly hatched chicks** were confirmed to be *Campylobacter*-free.
  - Birds were orally challenged with *C. jejuni* to establish infection.
  - Birds were assigned to two groups:
    - **Enrofloxacin-treated group:** received **50 mg/L enrofloxacin** in drinking water for **5 consecutive days**.
    - **Infected untreated control group:** received **sham treatment**.

- **Longitudinal sampling:**  
  Cecal contents were collected at **2, 7, 12, and 17 days post-treatment**

- **Analytical motivation:**  
  Because broiler chickens undergo rapid **gut maturation**, **host development**, and **microbial succession**, this dataset provides a useful setting for separating **antibiotic-associated microbial and metabolic changes** from background developmental variation.


### B. Data Preprocessing

We used `84_taxon_percent_report_tr_0.5pct.txt` as the microbiome relative abundance percentage table after filtering out species with less than 0.5% abundance in all samples. For the metabolome data, we used `IOST-02-21VW+ DATA TABLES, UPDATED.xlsx`, including the `Chemical Annotation`, `Sample Meta Data`, and `Log Transformed Data` sheets.

- The microbiome OTU data frame contains **84 samples**, **589 known species**, and **1 unclassified species**.
- The metabolome data frame contains **86 samples** and **793 known metabolites**.
- The metabolite group information data frame contains **793 metabolites** and **18 group annotation columns**, including **super pathway**, **sub pathway**, and **chemical name**.

After matching the microbiome and metabolome samples, the common samples included **6 baseline samples** and **73 other samples**. We used only the **73 non-baseline samples** for downstream analysis.

The final microbiome data frame contains **73 samples × 591 species**. One additional species was added so that the sum of each sample equals 1. The microbiome data were transformed back to **relative abundance** instead of percentage.

The sample information and feature group information are summarized below.

Note that the microbiome data are **relative abundances**, while the metabolome data are **log-transformed values**. For microbiome data, we applied the **centered log-ratio transformation** using `compositions::clr`, which replaces zeros with 1 and divides each relative abundance value by the geometric mean before taking the log. For metabolome data, we directly used the provided log-transformed values.

```{r helper_functions , eval = TRUE, include = FALSE}
make_time_label <- function(x, merge_late = FALSE) {
  x <- as.character(x)

  if (merge_late) {
    x[x %in% c("DPT-12", "DPT-17", "12", "17")] <- "Late (DPT-12/17)"
  }

  preferred <- c("DPT-2", "DPT-7", "DPT-12", "DPT-17",
                 "2", "7", "12", "17", "Late (DPT-12/17)")
  observed <- unique(x)
  lev <- c(preferred[preferred %in% observed],
           setdiff(observed, preferred))

  factor(x, levels = unique(lev))
}

safe_cosine <- function(x, y) {
  denom <- sqrt(sum(x^2)) * sqrt(sum(y^2))
  if (!is.finite(denom) || denom == 0) return(NA_real_)
  abs(sum(x * y) / denom)
}

get_sorted_loadings <- function(loading_matrix, feature_names, component = 1) {
  stopifnot(component <= ncol(loading_matrix))
  x <- loading_matrix[, component]
  names(x) <- feature_names
  x <- x[is.finite(x) & x != 0]
  x[order(abs(x), decreasing = TRUE)]
}

score_frame <- function(X, Y, u, v, metadata,
                        component = 1,
                        x_name = "score_X",
                        y_name = "score_Y",
                        merge_late = FALSE) {
  out <- data.frame(
    score_X = drop(X %*% u[, component, drop = FALSE]),
    score_Y = drop(Y %*% v[, component, drop = FALSE]),
    metadata,
    check.names = FALSE
  )
  names(out)[1:2] <- c(x_name, y_name)
  out$Time_raw <- as.character(out$Time)
  out$Time_plot <- make_time_label(out$Time, merge_late)
  out
}


plot_score_pair <- function(
    df,
    x,
    y,
    title,
    x_label,
    y_label,
    show_centroid_labels = TRUE
) {

  ## Check whether clustering assignments are available
  has_cluster <- "cluster" %in% names(df)

  ## Shape must be treated as a discrete variable
  if (has_cluster) {
    df <- df |>
      dplyr::mutate(cluster = as.factor(cluster))
  }

  ## Determine the temporal order using the unmerged time labels
  time_order <- c(
    "DPT-2", "DPT-7", "DPT-12", "DPT-17",
    "2", "7", "12", "17",
    "Late (DPT-12/17)"
  )

  ## Calculate centroids using the displayed time groups
  centroid_df <- df |>
    dplyr::group_by(Time_plot) |>
    dplyr::summarise(
      centroid_x = mean(.data[[x]], na.rm = TRUE),
      centroid_y = mean(.data[[y]], na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      Time_character = as.character(Time_plot),
      temporal_order = dplyr::case_when(
        Time_character %in% c("DPT-2", "2") ~ 1,
        Time_character %in% c("DPT-7", "7") ~ 2,
        Time_character %in% c("DPT-12", "12") ~ 3,
        Time_character %in% c("DPT-17", "17") ~ 4,
        Time_character == "Late (DPT-12/17)" ~ 3,
        TRUE ~ match(Time_character, time_order) + 10
      )
    ) |>
    dplyr::arrange(temporal_order)

  p <- ggplot2::ggplot(
    df,
    ggplot2::aes(
      x = .data[[x]],
      y = .data[[y]],
      color = Time_plot
    )
  )

  ## Individual samples
  if (has_cluster) {
    p <- p +
      ggplot2::geom_point(
        ggplot2::aes(shape = cluster),
        size = 2.8,
        alpha = 0.85
      )
  } else {
    p <- p +
      ggplot2::geom_point(
        shape = 16,
        size = 2.2,
        alpha = 0.85
      )
  }

  p <- p +
    ## Light gray dashed temporal trajectory
    ggplot2::geom_path(
      data = centroid_df,
      ggplot2::aes(
        x = centroid_x,
        y = centroid_y,
        group = 1
      ),
      inherit.aes = FALSE,
      color = "grey55",
      linewidth = 0.7,
      linetype = "dashed",
      alpha = 0.9,
      arrow = grid::arrow(
        type = "closed",
        length = grid::unit(0.12, "inches")
      )
    ) +

    ## Open-circle centroids
    ggplot2::geom_point(
      data = centroid_df,
      ggplot2::aes(
        x = centroid_x,
        y = centroid_y,
        fill = Time_plot
      ),
      inherit.aes = FALSE,
      shape = 21,
      size = 4.2,
      stroke = 1,
      color = "grey25",
      show.legend = FALSE
    ) +

    ggplot2::scale_color_brewer(
      palette = "Set1",
      drop = FALSE
    ) +
    
    ggplot2::scale_fill_brewer(
      palette = "Set1",
      drop = FALSE
    ) +

    ggplot2::labs(
      title = title,
      x = x_label,
      y = y_label,
      color = "Time",
      shape = "Cluster"
    ) +

    ggplot2::theme_bw(base_size = 12) +

    ggplot2::theme(
      panel.grid = ggplot2::element_blank(),
      text = ggplot2::element_text(size = 14),
      axis.title = ggplot2::element_text(size = 16),
      axis.text = ggplot2::element_text(size = 12),
      plot.title = ggplot2::element_text(face = "bold"),
      legend.position = "bottom"
    )

  return(p)
}

trajectory_correlation <- function(score_trt, score_ctrl,
                                   dpt_trt, dpt_ctrl) {
  trt_mean <- tapply(score_trt, dpt_trt, mean, na.rm = TRUE)
  ctrl_mean <- tapply(score_ctrl, dpt_ctrl, mean, na.rm = TRUE)

  common <- intersect(names(trt_mean), names(ctrl_mean))
  common <- common[order(suppressWarnings(as.numeric(gsub("[^0-9.]", "", common))))]

  if (length(common) < 3) {
    return(list(
      correlation = NA_real_,
      treatment_means = trt_mean,
      control_means = ctrl_mean,
      common_time = common
    ))
  }

  r <- cor(
    trt_mean[common],
    ctrl_mean[common],
    method = "pearson",
    use = "complete.obs"
  )

  list(
    correlation = unname(r),
    treatment_means = trt_mean,
    control_means = ctrl_mean,
    common_time = common
  )
}


get_top_loadings <- function(loading_matrix,
                             feature_names,
                             component = 1,
                             top_n = 10) {

  x <- loading_matrix[, component]
  names(x) <- feature_names

  x <- x[is.finite(x) & x != 0]
  x <- x[order(abs(x), decreasing = TRUE)]

  head(x, top_n)
}
make_temporal_matrix <- function(data_matrix,
                                 metadata,
                                 features,
                                 time_col = "Time") {

  sample_ids <- intersect(
    rownames(data_matrix),
    rownames(metadata)
  )

  df <- as.data.frame(
    data_matrix[sample_ids, features, drop = FALSE],
    check.names = FALSE
  ) |>
    rownames_to_column("SampleID")

  df$Time <- as.character(
    metadata[df$SampleID, time_col]
  )

  ## Standardize possible alternative time encodings
  df$Time <- dplyr::case_when(
    df$Time %in% c("2",  "DPT2",  "DPT 2")  ~ "DPT-2",
    df$Time %in% c("7",  "DPT7",  "DPT 7")  ~ "DPT-7",
    df$Time %in% c("12", "DPT12", "DPT 12") ~ "DPT-12",
    df$Time %in% c("17", "DPT17", "DPT 17") ~ "DPT-17",
    TRUE ~ df$Time
  )

  mean_matrix <- df |>
    filter(Time %in% time_order) |>
    select(-SampleID) |>
    group_by(Time) |>
    summarise(
      across(everything(), ~ mean(.x, na.rm = TRUE)),
      .groups = "drop"
    ) |>
    mutate(Time = factor(Time, levels = time_order)) |>
    arrange(Time) |>
    column_to_rownames("Time") |>
    as.matrix() |>
    t()

  ## Row-wise Z-score across time
  scaled_matrix <- t(scale(t(mean_matrix)))
  scaled_matrix[!is.finite(scaled_matrix)] <- 0

  scaled_matrix
}

fit_cpca <- function(target_data,
                     ancillary_data,
                     n_components = 2,
                     alpha = NULL) {

  target_data <- as.matrix(target_data)
  ancillary_data <- as.matrix(ancillary_data)

  cov_target <- stats::cov(target_data)
  cov_ancillary <- stats::cov(ancillary_data)

  ## Spectral-scale normalization within this modality
  if (is.null(alpha)) {
    lambda_target <- eigen(
      cov_target,
      symmetric = TRUE,
      only.values = TRUE
    )$values[1]

    lambda_ancillary <- eigen(
      cov_ancillary,
      symmetric = TRUE,
      only.values = TRUE
    )$values[1]

    if (!is.finite(lambda_ancillary) || lambda_ancillary <= 0) {
      stop("The leading ancillary eigenvalue must be positive.")
    }

    alpha <- lambda_target / lambda_ancillary
  }

  contrastive_covariance <- cov_target - alpha * cov_ancillary

  eig <- eigen(
    contrastive_covariance,
    symmetric = TRUE
  )

  ## eigen() returns values in decreasing order
  keep <- seq_len(min(n_components, ncol(eig$vectors)))

  target_centered <- sweep(
    target_data,
    2,
    colMeans(target_data),
    FUN = "-"
  )

  scores <- target_centered %*%
    eig$vectors[, keep, drop = FALSE]

  colnames(scores) <- paste0("cPC", keep)

  list(
    alpha = alpha,
    values = eig$values[keep],
    vectors = eig$vectors[, keep, drop = FALSE],
    scores = scores,
    contrastive_covariance = contrastive_covariance
  )
}


## Create the biological comparison label independently of merge_late
make_comparison_group <- function(x) {
  x <- as.character(x)

  dplyr::case_when(
    x %in% c("DPT-7", "7") ~ "DPT-7",
    x %in% c("DPT-12", "DPT-17", "12", "17") ~ "DPT-12/17",
    TRUE ~ NA_character_
  )
}
get_good_indices <- function(mat) {
  vars <- apply(mat, 2, var, na.rm = TRUE)
  nas  <- colSums(is.na(mat))
  return((vars > 1e-12) & (nas == 0))
}
```

```{r read in data without euka, include=FALSE, eval = TRUE}
root <- "~/Library/Mobile Documents/com~apple~CloudDocs/PhD/academic/2023 Spring/RA/ChickenData/"
mb <- read.table(file = paste0(root,'Shotgun metagenomics/84_taxon_percent_report_tr_0.5pct.txt'), sep = '\t', header = TRUE)
mtbgroup <- read.csv(file = paste0(root,"Metabolomics data from Metabolon Inc/mtbgroup.csv"), header = TRUE)
mtb <- read.csv(file = paste0(root,"Metabolomics data from Metabolon Inc/mtb.csv"), header = TRUE)
meta <- read.csv(file = paste0(root,"Metabolomics data from Metabolon Inc/meta.csv"), header = TRUE)
```

```{r data organization without euka, include=FALSE, eval = TRUE}
### create OTU data frame mb_dat (83 samples * 659 OTUs) 
OTU <- colnames(mb)[2:591]
mb_dat <- mb[,OTU]/100 # remove percentage
colnames(mb_dat) <- OTU
mb_dat$Others <- apply(mb_dat, 1, function(x){1-sum(x)}) # create others species to incorporate sum=1
rownames(mb_dat) <- sapply(mb$taxon_name,
                           function(x){strsplit(x,"-")[[1]][length(strsplit(x,"-")[[1]])]})


### create mtb data frame mtb_dat (86 samples * 1038 mtb) and mtb_group (1038 mtb * 18 group info)
mtb_dat <- left_join(meta, mtb, by = "PARENT_SAMPLE_NAME")
### only use common samples and non baseline samples: common 80, exclude 6 baseline, 73 samples will be used
common_samples <- intersect(mtb_dat$CLIENT_SAMPLE_ID, rownames(mb_dat))
nonB_common <- common_samples[sapply(common_samples, function(x){substr(x,1,1)})!="B"]
mtb_dat <- mtb_dat[mtb_dat$CLIENT_SAMPLE_ID %in% nonB_common,]
rownames(mtb_dat) <- mtb_dat$CLIENT_SAMPLE_ID
meta_dat <- mtb_dat[,1:5]
meta_dat$Time <- sapply(meta_dat$GROUP_ID, function(x){strsplit(x,"_")[[1]][2]})
meta_dat$Time <- factor(meta_dat$Time, levels=c("2d","7d","12d","17d"))
mtb_dat <- mtb_dat[,6:ncol(mtb_dat)]
colnames(mtb_dat) <- substr(colnames(mtb_dat),2,100)
colnames(mtb_dat) <- paste("mtb",colnames(mtb_dat), sep = "")
mtb_group <- mtbgroup
rownames(mtb_group) <- paste("mtb", mtbgroup$CHEM_ID, sep = "")
mtb_group[mtb_group$SUPER_PATHWAY == "", "SUPER_PATHWAY"] <- "Unknown"
mtb_group[mtb_group$SUB_PATHWAY == "", "SUB_PATHWAY"] <- "Unknown"
table(mtb_group$SUPER_PATHWAY)
table(mtb_group$SUB_PATHWAY)



# remove the mtbs and mbs with all 0/same values in 73 samples
mtb_dat <- mtb_dat[, apply(mtb_dat, 2, sd)!=0]
mtb_group <- mtb_group[colnames(mtb_dat),]
mb_dat <- mb_dat[nonB_common,]
mb_dat <- mb_dat[, apply(mb_dat, 2, sd) != 0]

### create mtb data frame mtb_dat_known (73 samples * 793 mtb) and mtb_group_known (793 mtb * 18 group info)
mtb_dat_known <- mtb_dat[,rownames(mtb_group)[mtb_group$TYPE=="NAMED"]]
mtb_group_known <- mtb_group[rownames(mtb_group)[mtb_group$TYPE=="NAMED"],]

# check data alignment
mb_dat <- mb_dat[rownames(mtb_dat),]
dim(mb_dat)
dim(mtb_dat)
dim(mtb_dat_known)
dim(meta_dat)
dim(mtb_group)
dim(mtb_group_known)
mean(rownames(mtb_group) == colnames(mtb_dat))
mean(rownames(mtb_dat) == rownames(meta_dat)) 
mean(rownames(mb_dat) == rownames(meta_dat)) 
mean(rownames(mtb_group_known) == colnames(mtb_dat_known))
mean(rownames(mtb_dat_known) == rownames(meta_dat))

colnames(mtb_dat_known) <- make.unique(mtb_group_known$SUB_PATHWAY)


## rename meta dat for visualization
meta_dat$TREATMENT <- factor(meta_dat$TREATMENT, levels=c("Control","Enrofloxacin"), labels=c("Control","ENRO"))
meta_dat$Time <- factor(meta_dat$Time, levels=c("2d","7d","12d","17d"), labels=c("DPT-2","DPT-7","DPT-12","DPT-17"))
## rename mb name for visualization
colnames(mb_dat) <- sapply(colnames(mb_dat), function(x){sub("\\.", " ", x)})

```

```{r data information print without euka , eval = TRUE}
print(knitr::kable(table(meta_dat$TREATMENT, meta_dat$Time), caption="Sample groups"))
print(knitr::kable(table(mtb_group_known$SUPER_PATHWAY), caption = "Superpathway summary of microbes",col.names = c("Superpathway","Counts")))
```


```{r s3CA_preprocessing, eval = TRUE}
X_all <- mtb_dat_known
Y_all <- t(apply(mb_dat, 1, compositions::clr))
target_ids <- rownames(meta_dat)[meta_dat$TREATMENT == "ENRO"]
backgr_ids <- rownames(meta_dat)[meta_dat$TREATMENT == "Control"]
Xt <- X_all[target_ids, ]; Xa <- X_all[backgr_ids, ]
Yt <- Y_all[target_ids, ]; Ya <- Y_all[backgr_ids, ]
meta_t <- meta_dat[target_ids, ]
meta_a <- meta_dat[backgr_ids, ]


keep_x <- get_good_indices(Xt) & get_good_indices(Xa)
keep_y <- get_good_indices(Yt) & get_good_indices(Ya)

Xt <- Xt[, keep_x]; Xa <- Xa[, keep_x]
Yt <- Yt[, keep_y]; Ya <- Ya[, keep_y]

Xt <- scale(Xt); Xa <- scale(Xa)
Yt <- scale(Yt); Ya <- scale(Ya)
```

### C. s3CA

We apply the s3CA algorithm. The parameter `eta` controls the strength of background removal.

*   **Standard CCA** (Target Only) captures the dominant variance in the treatment group.

*   **Standard CCA** (Ancillary Only) captures the dominant variance in the control group.

*   **Contrastive CCA** subtracts the background covariance to reveal unique structures.


```{r s3CA_analysis, eval = TRUE}
# eta is chosen by the silhouette - ARI criterion. The reference arrangement is the
# target-only leading pair with km_cluster = 2 (separating the acute DPT-2 samples), and
# the contrastive arrangement uses ckm_cluster = 3, matching the DPT-2 / DPT-7 / DPT-12-17
# partition reported below. Sparsity 0.2 is the value used for the analysis itself.
set.seed(111)
eta_tuning_broiler <- eta_tuning_bimodal(
  Xt = Xt, Yt = Yt, Xa = Xa, Ya = Ya,
  eta_range = seq(0.1, 3, 0.01),
  plot = TRUE,
  sparse_ctst = 0.2, sparse_trt = 0.2, sparse_ctrl = 0.2,
  reference = "target", num_comps = 1,
  km_cluster = 2, ckm_cluster = 3
)
eta_opt_broiler <- eta_tuning_broiler$eta_opt
cat("Selected eta:", eta_opt_broiler, "\n")

res_s3CA <- cCCA(Xa = Xa, Ya = Ya, Xt = Xt, Yt = Yt, eta = eta_opt_broiler,
                 sparse_ctst = 0.2, sparse_trt = 0.2, sparse_ctrl = 0.2)


CCA_ctrl <- res_s3CA$CCA_ctrl
CCA_trt  <- res_s3CA$CCA_trt
CCA_ctst <- res_s3CA$CCA_ctst
```




### D. Motivation: dominant acute structure and shared temporal structure

The leading treatment component is expected to emphasize the strongest treatment-specific mean structure. In this dataset, it primarily separates the acute DPT-2 samples from later time points. The second treatment component follows a smoother temporal trajectory. A similar trajectory appears in the first control component, motivating removal of longitudinal cross-modal structure represented in the ancillary population.

The score plots may optionally combine DPT-12 and DPT-17 into a single late label by setting `merge_late <- TRUE` in the setup chunk. The trajectory correlations below always use the original four time points.


```{r, eval = TRUE}
merge_late <- FALSE
```

```{r motivation_scores, eval = TRUE, fig.width=15, fig.height=5, out.width="100%"}
df_trt_c1 <- score_frame(
  Xt, Yt, CCA_trt$u, CCA_trt$v, meta_t,
  component = 1,
  x_name = "CCX1", y_name = "CCY1",
  merge_late = merge_late
)
df_trt_c1$cluster <- spec_clust(df_trt_c1[,1:2], k=2, group = df_trt_c1$Time)$cluster

df_trt_c2 <- score_frame(
  Xt, Yt, CCA_trt$u, CCA_trt$v, meta_t,
  component = 2,
  x_name = "CCX2", y_name = "CCY2",
  merge_late = merge_late
)
df_trt_c2$cluster <- spec_clust(df_trt_c2[,1:2], k=3, group = df_trt_c2$Time)$cluster


df_ctrl_c1 <- score_frame(
  Xa, Ya, CCA_ctrl$u, CCA_ctrl$v, meta_a,
  component = 1,
  x_name = "CCX1", y_name = "CCY1",
  merge_late = merge_late
)
df_ctrl_c1$cluster <- spec_clust(df_ctrl_c1[,1:2], k=3, group = df_ctrl_c1$Time)$cluster


p_trt_c1 <- plot_score_pair(
  df_trt_c1, "CCX1", "CCY1",
  "Treatment pair 1",
  "Metabolome score",
  "Microbiome score"
)

p_trt_c2 <- plot_score_pair(
  df_trt_c2, "CCX2", "CCY2",
  "Treatment pair 2",
  "Metabolome score",
  "Microbiome score"
)

p_ctrl_c1 <- plot_score_pair(
  df_ctrl_c1, "CCX1", "CCY1",
  "Control pair 1",
   "Metabolome score",
  "Microbiome score"
)

# p_motivation <- ggarrange(
#   p_trt_c1, p_trt_c2, p_ctrl_c1,
#   nrow = 1,
#   common.legend = TRUE,
#   legend = "bottom",
#   labels = "AUTO"
# )

# print(p_motivation)

## Extract the complete legend from a plot that contains all three clusters
shared_legend <- cowplot::get_legend(
  p_trt_c2 +
    ggplot2::theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.margin = ggplot2::margin(
        t = 0,
        r = 0,
        b = 0,
        l = 0
      )
    )
)

## Arrange plots in the original order without individual legends
plot_row <- ggpubr::ggarrange(
  p_trt_c1 + ggplot2::theme(legend.position = "none"),
  p_trt_c2 + ggplot2::theme(legend.position = "none"),
  p_ctrl_c1 + ggplot2::theme(legend.position = "none"),
  ncol = 3,
  nrow = 1,
  labels = c("A", "B", "C"),
  align = "hv"
)

## Add the shared legend below
final_plot <- cowplot::plot_grid(
  plot_row,
  shared_legend,
  ncol = 1,
  rel_heights = c(1, 0.10)
)

final_plot

# motivation_file <- file.path(
#   "../inst/SavedFigures",
#   if (merge_late) {
#     "Chicken_motivation_scores_late_merged.pdf"
#   } else {
#     "Chicken_motivation_scores.pdf"
#   }
# )

# ggsave(
#   motivation_file,
#   plot = final_plot,
#   width = 15,
#   height = 5
# )

# cat("Saved motivation figure to:", motivation_file, "\n")
```

#### D-1 Temporal-score concordance

The temporal concordance is summarized using Pearson correlations between the four DPT-specific score centroids from treatment component 2 and control component 1. Because the comparison uses only four time points, the correlations are descriptive and should be interpreted together with the score plots.

```{r temporal_score_correlation, eval = TRUE}
scoreX_trt  <- Xt %*% as.matrix(CCA_trt$u)
scoreY_trt  <- Yt %*% as.matrix(CCA_trt$v)
scoreX_ctrl <- Xa %*% as.matrix(CCA_ctrl$u)
scoreY_ctrl <- Ya %*% as.matrix(CCA_ctrl$v)

dpt_trt  <- as.character(meta_t$Time)
dpt_ctrl <- as.character(meta_a$Time)

traj_X <- trajectory_correlation(
  scoreX_trt[, 2],
  scoreX_ctrl[, 1],
  dpt_trt,
  dpt_ctrl
)

traj_Y <- trajectory_correlation(
  scoreY_trt[, 2],
  scoreY_ctrl[, 1],
  dpt_trt,
  dpt_ctrl
)

temporal_correlations <- data.frame(
  modality = c("Metabolome", "Microbiome"),
  treatment_component = 2,
  control_component = 1,
  pearson_r = c(traj_X$correlation, traj_Y$correlation),
  n_timepoints = c(length(traj_X$common_time), length(traj_Y$common_time))
)

print(temporal_correlations)

cat(sprintf(
  paste0(
    "Temporal centroid correlation: metabolome r = %.3f; ",
    "microbiome r = %.3f.\n"
  ),
  traj_X$correlation,
  traj_Y$correlation
))
```

**Interpretation.** The first treatment pair is dominated by the acute DPT-2 separation, whereas the second treatment pair shows a temporal organization strongly concordant with the leading control pair. This supports the presence of a shared sample-level temporal pattern at different spectral ranks. It does not require the sparse loading vectors to be identical.

#### D-2 Supplementary loading-alignment heatmap

Sparse solutions may use different correlated features to represent similar sample trajectories. We therefore treat loading alignment as a supplementary diagnostic rather than the primary evidence for temporal concordance.

```{r supp_alignment_heatmap, eval = TRUE}
u_ctrl <- CCA_ctrl$u
v_ctrl <- CCA_ctrl$v
u_trt  <- CCA_trt$u
v_trt  <- CCA_trt$v

K <- min(5, ncol(u_trt), ncol(v_trt))

alignment_df <- lapply(seq_len(K), function(k) {
  cos_u <- safe_cosine(u_ctrl[, 1], u_trt[, k])
  cos_v <- safe_cosine(v_ctrl[, 1], v_trt[, k])

  data.frame(
    treatment_component = paste0("Treatment C", k),
    Metabolome = cos_u,
    Microbiome = cos_v,
    Joint = cos_u * cos_v
  )
}) |>
  bind_rows()

alignment_long <- alignment_df |>
  pivot_longer(
    cols = c(Metabolome, Microbiome, Joint),
    names_to = "alignment_type",
    values_to = "cosine_similarity"
  ) |>
  mutate(
    treatment_component = factor(
      treatment_component,
      levels = paste0("Treatment C", seq_len(K))
    ),
    alignment_type = factor(
      alignment_type,
      levels = c("Metabolome", "Microbiome", "Joint")
    )
  )

p_alignment <- ggplot(
  alignment_long,
  aes(
    x = treatment_component,
    y = alignment_type,
    fill = cosine_similarity
  )
) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(
    aes(label = sprintf("%.2f", cosine_similarity)),
    size = 4
  ) +
  scale_fill_gradient(
    low = "white",
    high = "steelblue",
    limits = c(0, 1),
    name = "Absolute\ncosine similarity"
  ) +
  labs(
    x = "Treatment component",
    y = NULL,
    title = "Alignment with control component 1",
    subtitle = "Cosine similarity of sparse loading vectors"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )

print(p_alignment)

# alignment_file <- file.path(
#   "../inst/SavedFigures",
#   "Chicken_supp_loading_alignment_heatmap.pdf"
# )

# ggsave(
#   alignment_file,
#   plot = p_alignment,
#   width = 10,
#   height = 5
# )


# cat("Saved supplementary alignment heatmap to:", alignment_file, "\n")
```

#### D-3 The temporal variation is shared between the two groups

Model (\ref{eq:bi}) represents the temporal variation as a latent factor with a common pair of
loading vectors and group-specific strength. Development and post-infection recovery are
continuous processes sampled at four days, which is why a factor rather than four unrelated
nuisance clusters is used; the four day-means are not collinear, so the choice rests on the
biology rather than on the rank of the observed structure.

The check below asks whether the treated birds follow the same temporal pattern as the
untreated birds. Definitions used throughout this subsection:

* **Score pair.** For a loading pair $(\hat\bu,\hat\bv)$, the two-column matrix
  $(\bX\hat\bu,\ \bY\hat\bv)$ — one score per modality. 
* **Time profile.** The mean of one modality's score within each of the four days: four numbers
  per group per modality. Reported per modality, because the two modalities behave differently
  here and a combined score would average them together.
* **Shape** $=$ Pearson correlation between the treated and untreated four-number profiles, with
  *both groups projected onto the same loadings*. Because both profiles use the same loading
  vector, the arbitrary sign of a singular vector flips both together and the correlation is
  unchanged, so its sign is meaningful: $+1$ means the two groups rise and fall together over
  time, $-1$ means they move oppositely, $0$ means unrelated.
* **Amplitude** $=$ standard deviation of one group's four day-means. Model (\ref{eq:bi}) allows
  the groups to share the shape while differing in strength ($\tau_\tt$ versus $\tau_\aa$), so
  shape and amplitude are reported separately.
* **Data layer.** `Xt`, `Xa`, `Yt`, `Ya` are centered and scaled *within each group*, so these
  values are relative to that group's own average. They may be compared in shape across days,
  not in level between groups. All loadings come from the sparse fit (`sumabs = 0.2`).
* Profiles use only four points, so shape is reported with a bootstrap interval obtained by
  resampling samples within each day.

```{r broiler_shared_trajectory, eval = TRUE}
dpt_levels <- c("DPT-2", "DPT-7", "DPT-12", "DPT-17")
time_t <- factor(as.character(meta_t$Time), levels = dpt_levels)
time_a <- factor(as.character(meta_a$Time), levels = dpt_levels)

## every loading pair below comes from the sparse fit computed in `s3CA_analysis`
loadings <- list(
  "Untreated pair 1"   = list(u = CCA_ctrl$u[, 1], v = CCA_ctrl$v[, 1]),
  "Treatment pair 1"   = list(u = CCA_trt$u[, 1],  v = CCA_trt$v[, 1]),
  "Treatment pair 2"   = list(u = CCA_trt$u[, 2],  v = CCA_trt$v[, 2]),
  "Contrastive pair 1" = list(u = CCA_ctst$u[, 1], v = CCA_ctst$v[, 1])
)

## day-means of one modality's score, for each group, on the SAME loading vector
profiles <- function(w, Dt, Da) {
  list(treated   = tapply(drop(Dt %*% w), time_t, mean)[dpt_levels],
       untreated = tapply(drop(Da %*% w), time_a, mean)[dpt_levels])
}

set.seed(2024)
boot_shape <- function(w, Dt, Da) {
  st <- drop(Dt %*% w); sa <- drop(Da %*% w)
  r <- replicate(2000, {
    it <- unlist(lapply(split(seq_along(time_t), time_t), function(i) sample(i, length(i), TRUE)))
    ia <- unlist(lapply(split(seq_along(time_a), time_a), function(i) sample(i, length(i), TRUE)))
    suppressWarnings(cor(tapply(st[it], time_t[it], mean)[dpt_levels],
                         tapply(sa[ia], time_a[ia], mean)[dpt_levels]))
  })
  quantile(r, c(0.025, 0.975), na.rm = TRUE)
}

shared_table <- do.call(rbind, lapply(names(loadings), function(nm) {
  u <- loadings[[nm]]$u; v <- loadings[[nm]]$v
  pm <- profiles(u, Xt, Xa)      # metabolome
  pb <- profiles(v, Yt, Ya)      # microbiome
  cim <- boot_shape(u, Xt, Xa); cib <- boot_shape(v, Yt, Ya)
  data.frame(
    loadings       = nm,
    cor_mtb      = cor(pm$treated, pm$untreated),
    mtb_CI         = sprintf("[%.2f, %.2f]", cim[1], cim[2]),
    cor_mb       = cor(pb$treated, pb$untreated),
    mb_CI          = sprintf("[%.2f, %.2f]", cib[1], cib[2]),
    corsd_mb_treated = sd(pb$treated),
    corsd_mb_untrt   = sd(pb$untreated)
  )
}))
print(knitr::kable(shared_table, digits = 3, row.names = FALSE, format = "markdown",
      caption = "Do the two groups follow the same time profile along the same loadings? Reported separately for each modality."))
## Reference distribution. A shape correlation is only interpretable against what an
## uninformative direction would give, and that baseline differs between the modalities.
## Here 300 random loading vectors are drawn per modality and put through the same
## projection, day-averaging and correlation.
shape_of <- function(w, Dt, Da) {
  pr <- profiles(w, Dt, Da)
  suppressWarnings(cor(pr$treated, pr$untreated))
}
set.seed(1)
null_mtb <- replicate(300, shape_of(rnorm(ncol(Xt)), Xt, Xa))
null_mb  <- replicate(300, replicate(1, shape_of(rnorm(ncol(Yt)), Yt, Ya)))

## The baseline is set by how many individual features already agree between the groups:
## a random weighted average of features that mostly agree will itself agree.
feature_agreement <- function(Dt, Da) {
  day_profile <- function(D, f) {
    M <- t(sapply(seq_len(ncol(D)), function(j) tapply(D[, j], f, mean)[dpt_levels]))
    t(scale(t(M), scale = FALSE))
  }
  Mt <- day_profile(Dt, time_t); Ma <- day_profile(Da, time_a)
  r  <- sapply(seq_len(nrow(Mt)), function(i) suppressWarnings(cor(Mt[i, ], Ma[i, ])))
  c(median = median(r, na.rm = TRUE), pct_positive = 100 * mean(r > 0, na.rm = TRUE))
}
ag_mtb <- feature_agreement(Xt, Xa); ag_mb <- feature_agreement(Yt, Ya)

null_table <- data.frame(
  modality          = c("Metabolome", "Microbiome"),
  null_median_shape = c(median(null_mtb, na.rm = TRUE), median(null_mb, na.rm = TRUE)),
  null_pct_above_07 = c(100 * mean(null_mtb > 0.7, na.rm = TRUE),
                        100 * mean(null_mb  > 0.7, na.rm = TRUE)),
  feature_median_r  = c(ag_mtb["median"], ag_mb["median"]),
  feature_pct_agree = c(ag_mtb["pct_positive"], ag_mb["pct_positive"])
)
print(knitr::kable(null_table, digits = 3, row.names = FALSE, format = "markdown",
      caption = "Reference distribution. Columns 2-3: shape correlation from 300 random loading vectors. Columns 4-5: correlation between the two groups' four-day profiles for each feature separately, and the percentage of features whose profiles agree in direction."))

```

The metabolome profile is similar between the groups under every loading pair, so on its own it
does not identify which component carries shared variation. The microbiome profile discriminates:
only the second treatment pair shows the two groups moving together, which identifies it as the
component carrying the shared temporal factor.

#### D-4 Is the DPT-7 distinction specific to the treated birds?

Two comparisons must be kept apart.

* **DPT-2 versus the later days** is *shared*: the acute phase differs from later days in every
  bird. A component separating those is not evidence of a treatment effect.
* **DPT-7 versus DPT-12/17, among post-acute samples only**, is the finding. For it to be
  treatment-specific it must be strong in the treated birds and weak in the untreated birds
  *along the same loadings*.

**ARI** compares a k-means partition of the two-column score pair $(\bX\hat\bu, \bY\hat\bv)$ with
the stated day grouping. It has expected value $0$ for unrelated partitions and $1$ for perfect
agreement, and is computed separately within each group against that group's own day labels. Day
labels are never used in fitting, so a high ARI is not circular.

```{r broiler_attribution_table, eval = TRUE}
group3 <- function(x) ifelse(as.character(x) %in% c("DPT-12", "DPT-17"),
                             "DPT-12/17", as.character(x))
later_t <- as.character(time_t) != "DPT-2"
later_a <- as.character(time_a) != "DPT-2"

attribution_table <- do.call(rbind, lapply(names(loadings), function(nm) {
  u <- loadings[[nm]]$u; v <- loadings[[nm]]$v
  pair_t <- cbind(Xt %*% u, Yt %*% v)          # two columns, as everywhere else
  pair_a <- cbind(Xa %*% u, Ya %*% v)
  data.frame(
    loadings       = nm,
    dpt7_treated   = spec_clust(pair_t[later_t, ], k = 2,
                                group = group3(time_t[later_t]), seed = 1)$ari,
    dpt7_untreated = spec_clust(pair_a[later_a, ], k = 2,
                                group = group3(time_a[later_a]), seed = 1)$ari
  )
}))
attribution_table$gap <- attribution_table$dpt7_treated - attribution_table$dpt7_untreated
print(knitr::kable(attribution_table, digits = 3, row.names = FALSE, format = "markdown",
      caption = "Separation of DPT-7 from DPT-12/17 within each group. A large gap means the distinction exists in the treated birds but not the untreated ones."))

      
```

The second treatment pair separates the later days in both groups, so a treatment-only analysis
cannot attribute that separation to the treatment. The leading contrastive pair separates them in
the treated birds only.

#### D-5 What the contrastive component describes: a slowed transition

Chunk D-3 showed that the two groups' metabolome profiles are highly correlated along the
contrastive loadings (0.921). That says they change in the same *direction*. It does not say they
change at the same *time*, because a correlation is invariant to how each profile is centred and
scaled. The statistic below isolates the timing.

**Progression fraction** at day $t$, computed per group as
$(\bar s(t)-\bar s(\text{DPT-2}))/(\bar s(\text{DPT-17})-\bar s(\text{DPT-2}))$, where $\bar s(t)$
is that group's mean metabolome score on day $t$. It equals 0 at DPT-2 and 1 at DPT-17 by
construction, so only the intermediate days carry information. Multiplying the score by any
constant leaves it unchanged, so unlike the clustering ARI it does not depend on how the score
columns are scaled.

**Why this explains the clustering.** A day forms its own cluster only if it is far from *both*
endpoints; on a 0-1 scale, 0.5 is maximally distinct and either extreme is not distinct at all.
The `distance_to_nearer_end` column below is $\min(p, 1-p)$ and is the quantity that matters.

**Metabolome only.** No progression fraction is reported for the microbiome, because the two
groups move in opposite directions there (shape correlation $-0.928$). The denominator
$\bar s(\text{DPT-17})-\bar s(\text{DPT-2})$ would then be positive in one group and negative in
the other, both groups would rescale to 0 at DPT-2 and 1 at DPT-17, and the reversal would be
hidden rather than measured. The microbiome is shown instead in the two-group heatmap of
Section F, where the opposing directions are visible directly.

```{r broiler_delayed_transition, eval = TRUE}
u_ctst <- CCA_ctst$u[, 1]; v_ctst <- CCA_ctst$v[, 1]

prog <- function(score, tvec) {
  m <- tapply(score, tvec, mean)[dpt_levels]
  (m - m["DPT-2"]) / (m["DPT-17"] - m["DPT-2"])
}
p_t <- prog(drop(Xt %*% u_ctst), time_t)
p_a <- prog(drop(Xa %*% u_ctst), time_a)

progression <- data.frame(
  group                   = c("Treated", "Untreated"),
  rbind(round(p_t, 3), round(p_a, 3)),
  distance_to_nearer_end  = round(c(min(p_t["DPT-7"], 1 - p_t["DPT-7"]),
                                    min(p_a["DPT-7"], 1 - p_a["DPT-7"])), 3),
  check.names = FALSE
)
print(knitr::kable(progression, row.names = FALSE, format = "markdown",
      caption = "Fraction of each group's own DPT-2 to DPT-17 metabolome change completed by each day. The last column is the DPT-7 distance to the nearer endpoint: larger means DPT-7 is more separable from both the acute and the later days."))

## The same question at the feature level for the microbiome: on each selected taxon, is DPT-7
## nearer to DPT-2 or to DPT-12 within each group? Values are per-group standardized, so they
## are comparable in shape across days but not in level between groups.
sel_mb <- {i <- which(v_ctst != 0); i[order(abs(v_ctst[i]), decreasing = TRUE)][1:10]}
day_mean <- function(D, fac, idx) sapply(dpt_levels,
  function(d) colMeans(D[fac == d, idx, drop = FALSE]))
Mt_mb <- day_mean(Yt, time_t, sel_mb); Ma_mb <- day_mean(Ya, time_a, sel_mb)

microbe_tbl <- data.frame(
  microbe = substr(colnames(Yt)[sel_mb], 1, 32),
  loading = round(v_ctst[sel_mb], 3),
  round(Mt_mb, 2), round(Ma_mb, 2),
  ## which day is DPT-7 nearer to, within each group?
  trt_DPT7_nearer = ifelse(abs(Mt_mb[, "DPT-7"] - Mt_mb[, "DPT-2"]) <
                           abs(Mt_mb[, "DPT-7"] - Mt_mb[, "DPT-12"]), "DPT-2", "DPT-12"),
  ctl_DPT7_nearer = ifelse(abs(Ma_mb[, "DPT-7"] - Ma_mb[, "DPT-2"]) <
                           abs(Ma_mb[, "DPT-7"] - Ma_mb[, "DPT-12"]), "DPT-2", "DPT-12"),
  check.names = FALSE
)
names(microbe_tbl)[3:6] <- paste0("trt_", dpt_levels)
names(microbe_tbl)[7:10] <- paste0("ctl_", dpt_levels)
print(knitr::kable(microbe_tbl, row.names = FALSE, format = "markdown",
      caption = "Selected microbial features. The last two columns ask, within each group, whether DPT-7 sits nearer to the acute day or to DPT-12."))
```

In the treated birds DPT-7 sits near the middle of the trajectory and is therefore far from both
ends; in the untreated birds it has already arrived close to the later days. This is why DPT-7
forms a separate cluster in one group and not the other. The antibiotic does not change where the
metabolome ends up, only how quickly it gets there.


### E. Contrastive target structure

The contrastive analysis retains the large acute DPT-2 structure while reducing cross-modal variation represented in the untreated controls. The leading contrastive score pair can therefore place greater emphasis on post-acute treatment structure, including the separation of DPT-7 from DPT-12/17.

```{r contrastive_score_plot, eval = TRUE}
df_ctst_c1 <- score_frame(
  Xt, Yt, CCA_ctst$u, CCA_ctst$v, meta_t,
  component = 1,
  x_name = "cCCX1", y_name = "cCCY1",
  merge_late = merge_late
)
df_ctst_c1$cluster <- spec_clust(df_ctst_c1[,1:2], k=3, group = df_ctst_c1$Time)$cluster

p_ctst <- plot_score_pair(
  df_ctst_c1,
  "cCCX1", "cCCY1",
  "",
  "Contrastive metabolome score",
  "Contrastive microbiome score"
)

print(p_ctst)

# contrastive_file <- file.path(
#   "../inst/SavedFigures",
#   if (merge_late) {
#     "Chicken_contrastive_score_late_merged.pdf"
#   } else {
#     "Chicken_contrastive_score.pdf"
#   }
# )

# ggsave(
#   contrastive_file,
#   plot = p_ctst,
#   width = 7.5,
#   height = 6
# )

# cat("Saved contrastive score plot to:", contrastive_file, "\n")
```

**Interpretation.** The contrastive score plot should be read as a joint two-dimensional representation: the horizontal coordinate is the metabolome score and the vertical coordinate is the microbiome score from the same singular-vector pair. Distinct time-specific centroids therefore indicate coordinated differences across both modalities.

### F. Selected features and temporal mean heatmaps

We next inspect features selected by the first contrastive component. Features are ranked by the absolute loading magnitude, but the signed loading is retained in the output table.

```{r selected_contrastive_features, eval = TRUE}
mtb_loadings_ctst <- get_sorted_loadings(
  CCA_ctst$u,
  colnames(Xt),
  component = 1
)

mb_loadings_ctst <- get_sorted_loadings(
  CCA_ctst$v,
  colnames(Yt),
  component = 1
)

top_n_mtb <- min(10, length(mtb_loadings_ctst))
top_n_mb  <- min(10, length(mb_loadings_ctst))

selected_mtb <- head(mtb_loadings_ctst, top_n_mtb)
selected_mb  <- head(mb_loadings_ctst, top_n_mb)

cat("Selected metabolites from contrastive component 1:\n")
print(selected_mtb)

cat("\nSelected microbiome features from contrastive component 1:\n")
print(selected_mb)

selected_feature_table <- bind_rows(
  data.frame(
    feature = names(selected_mtb),
    loading = as.numeric(selected_mtb),
    modality = "Metabolome"
  ),
  data.frame(
    feature = names(selected_mb),
    loading = as.numeric(selected_mb),
    modality = "Microbiome"
  )
)

# write.csv(
#   selected_feature_table,
#   file.path("../inst/SavedFigures", "Chicken_contrastive_selected_features.csv"),
#   row.names = FALSE
# )
```

The heatmaps display the mean abundance of each selected feature at each of the four original time points. Each row is standardized across time, so the colors describe the temporal pattern of a feature rather than its absolute abundance relative to other features.



```{r ctst_feature_heatmaps, eval = TRUE, fig.width=14, fig.height=6.5, out.width="100%"}
## Contrastive C1 temporal feature heatmaps
top_n_heatmap <- 10
time_order <- c("DPT-2", "DPT-7", "DPT-12", "DPT-17")

mtb_loadings_ctst <- get_top_loadings(
  CCA_ctst$u,
  colnames(Xt),
  component = 1,
  top_n = top_n_heatmap
)

mb_loadings_ctst <- get_top_loadings(
  CCA_ctst$v,
  colnames(Yt),
  component = 1,
  top_n = top_n_heatmap
)



mat_mtb_ctst <- make_temporal_matrix(
  Xt,
  meta_t,
  names(mtb_loadings_ctst)
)

mat_mb_ctst <- make_temporal_matrix(
  Yt,
  meta_t,
  names(mb_loadings_ctst)
)


## Wrap long feature names onto multiple lines
wrap_labels <- function(x, width = 28) {
  vapply(
    x,
    function(label) {
      paste(strwrap(label, width = width), collapse = "\n")
    },
    character(1)
  )
}

mtb_labels <- wrap_labels(rownames(mat_mtb_ctst), width = 28)
mb_labels  <- wrap_labels(rownames(mat_mb_ctst),  width = 28)

ph_mtb <- pheatmap(
  mat_mtb_ctst,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  display_numbers = FALSE,
  color = colorRampPalette(c("white", "navy"))(50),
  labels_row = mtb_labels,
    main = "Contrastive C1 – Metabolome",
  fontsize_row = 11,
  fontsize_col = 10,
  angle_col = 0,
  border_color = NA,
  legend = TRUE,
  silent = TRUE
)

ph_mb <- pheatmap(
  mat_mb_ctst,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  display_numbers = FALSE,
  color = colorRampPalette(c("white", "firebrick3"))(50),
  labels_row = mb_labels,
  main = "Contrastive C1 – Microbiome",
  fontsize_row = 11,
  fontsize_col = 10,
  angle_col = 0,
  border_color = NA,
  legend = TRUE,
  silent = TRUE
)

combined_heatmap <- arrangeGrob(
  ph_mtb$gtable,
  ph_mb$gtable,
  ncol = 2
)

grid.newpage()
grid.draw(combined_heatmap)

# combined_file <- file.path(
#   "../inst/SavedFigures",
#   "Chicken_contrastive_feature_heatmaps.pdf"
# )

# pdf(combined_file, width = 14, height = 6.5)
# grid.newpage()
# grid.draw(combined_heatmap)
# dev.off()

# cat("Saved combined contrastive heatmap to:", combined_file, "\n")
```


**Interpretation.** These heatmaps describe how the selected features contribute to the contrastive temporal organization. They should not be interpreted as causal microbe–metabolite interactions. The main result is that the selected metabolites and microbial features jointly distinguish the acute, intermediate, and later treatment states.

```{r broiler_two_group_heatmap, eval = TRUE, fig.width=14, fig.height=6.5, out.width="100%"}
## Same features as the contrastive heatmap, but shown in BOTH groups. Rows are z-scored
## across all eight columns, so each row describes a temporal pattern and the two blocks are
## directly comparable; absolute levels are not, because the two groups were scaled separately.
time_order <- c("DPT-2", "DPT-7", "DPT-12", "DPT-17")

## day means for one group, unscaled
day_means <- function(D, meta, features) {
  ids <- intersect(rownames(D), rownames(meta))
  tm  <- factor(as.character(meta[ids, "Time"]), levels = time_order)
  out <- sapply(time_order, function(d)
    colMeans(D[ids[tm == d], features, drop = FALSE], na.rm = TRUE))
  matrix(out, nrow = length(features), dimnames = list(features, time_order))
}

## treated and untreated blocks side by side, row z-scored across all eight columns
make_temporal_matrix_2group <- function(D_t, meta_t, D_a, meta_a, features) {
  mat <- cbind(day_means(D_t, meta_t, features), day_means(D_a, meta_a, features))
  colnames(mat) <- c(paste0("Treated\n",   time_order),
                   paste0("Untreated\n", time_order))
  m <- t(scale(t(mat)))
  m[!is.finite(m)] <- 0
  m
}

mat_mtb_2g <- make_temporal_matrix_2group(
  Xt, meta_t, Xa, meta_a, names(mtb_loadings_ctst))
mat_mb_2g <- make_temporal_matrix_2group(
  Yt, meta_t, Ya, meta_a, names(mb_loadings_ctst))

ph_mtb_2g <- pheatmap(
  mat_mtb_2g, cluster_rows = TRUE, cluster_cols = FALSE, gaps_col = 4,
  color = colorRampPalette(c("white", "navy"))(50),
  labels_row = wrap_labels(rownames(mat_mtb_2g), width = 28),
  main = "Contrastive C1 - Metabolome",
  fontsize_row = 11, fontsize_col = 10, angle_col = 0,
  border_color = NA, legend = TRUE, silent = TRUE
)

ph_mb_2g <- pheatmap(
  mat_mb_2g, cluster_rows = TRUE, cluster_cols = FALSE, gaps_col = 4,
  color = colorRampPalette(c("white", "firebrick3"))(50),
  labels_row = wrap_labels(rownames(mat_mb_2g), width = 28),
  main = "Contrastive C1 - Microbiome",
  fontsize_row = 11, fontsize_col = 10, angle_col = 0,
  border_color = NA, legend = TRUE, silent = TRUE
)

combined_2g <- arrangeGrob(ph_mtb_2g$gtable, ph_mb_2g$gtable, ncol = 2)
grid.newpage()
grid.draw(combined_2g)

## how far has each group moved by DPT-7, relative to its own DPT-2 state?
cat("\nmean |treated DPT-7 - untreated DPT-2| :",
    sprintf("%.3f", mean(abs(mat_mtb_2g[, 2] - mat_mtb_2g[, 5]))), "\n")
cat("mean |treated DPT-7 - untreated DPT-7| :",
    sprintf("%.3f", mean(abs(mat_mtb_2g[, 2] - mat_mtb_2g[, 6]))), "\n")

# pdf(file.path("../inst/SavedFigures", "Chicken_contrastive_feature_heatmaps_2group.pdf"),
#     width = 14, height = 6.5)
# grid.newpage(); grid.draw(combined_2g); dev.off()
```

**Interpretation.** Read each panel by comparing the two blocks column by column. In the
metabolome the treated block turns over one column later than the untreated block: the top row
group is dark at both `Trt DPT-2` and `Trt DPT-7` but only at `Ctl DPT-2`, and the bottom row
group shows the same offset inverted. The slowdown is partial rather than a shift by a full
sampling interval, since the treated DPT-7 profile differs from the untreated DPT-2 profile by
1.02 on this scale against 0.86 from the untreated DPT-7 profile.

The microbiome panel contains two kinds of row. Four *Citrobacter* taxa show the same one-column
offset as the metabolome, pale at `Trt DPT-2` and `Trt DPT-7` and dark from `Trt DPT-12`, while
in the untreated block only `Ctl DPT-2` is distinct; the table in D-5 records this as DPT-7 lying
nearer the acute day in the treated birds and nearer DPT-12 in the untreated ones. These are the
rows that make the microbiome contribute to the DPT-7 separation. The remaining taxa, including
the normal-flora species, change in opposite directions in the two groups and do not show the
offset.

```{r trt_feature_heatmaps, eval = TRUE, fig.width=14, fig.height=6.5, out.width="100%"}
## Treatment-only component 1
mtb_loadings_trt <- get_top_loadings(
  CCA_trt$u,
  colnames(Xt),
  component = 1,
  top_n = top_n_heatmap
)

mb_loadings_trt <- get_top_loadings(
  CCA_trt$v,
  colnames(Yt),
  component = 1,
  top_n = top_n_heatmap
)

mat_mtb_trt <- make_temporal_matrix(
  Xt,
  meta_t,
  names(mtb_loadings_trt)
)

mat_mb_trt <- make_temporal_matrix(
  Yt,
  meta_t,
  names(mb_loadings_trt)
)

mtb_labels_trt <- wrap_labels(
  rownames(mat_mtb_trt),
  width = 28
)

mb_labels_trt <- wrap_labels(
  rownames(mat_mb_trt),
  width = 28
)

ph_mtb_trt <- pheatmap(
  mat_mtb_trt,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  display_numbers = FALSE,
  color = colorRampPalette(c("white", "navy"))(50),
  labels_row = mtb_labels_trt,
  main = "Treatment C1 – Metabolome",
  fontsize_row = 11,
  fontsize_col = 10,
  angle_col = 0,
  border_color = NA,
  legend = TRUE,
  silent = TRUE
)

ph_mb_trt <- pheatmap(
  mat_mb_trt,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  display_numbers = FALSE,
  color = colorRampPalette(c("white", "firebrick3"))(50),
  labels_row = mb_labels_trt,
  main = "Treatment C1 – Microbiome",
  fontsize_row = 11,
  fontsize_col = 10,
  angle_col = 0,
  border_color = NA,
  legend = TRUE,
  silent = TRUE
)

combined_heatmap_trt <- arrangeGrob(
  ph_mtb_trt$gtable,
  ph_mb_trt$gtable,
  ncol = 2
)

grid.newpage()
grid.draw(combined_heatmap_trt)

# treatment_file <- file.path(
#   "../inst/SavedFigures",
#   "Chicken_treatment_feature_heatmaps.pdf"
# )
#
# pdf(treatment_file, width = 14, height = 6.5)
# grid.newpage()
# grid.draw(combined_heatmap_trt)
# dev.off()
```



```{r ctrl_feature_heatmaps, eval = TRUE, fig.width=14, fig.height=6.5, out.width="100%"}
## Control-only component 1
mtb_loadings_ctrl <- get_top_loadings(
  CCA_ctrl$u,
  colnames(Xa),
  component = 1,
  top_n = top_n_heatmap
)

mb_loadings_ctrl <- get_top_loadings(
  CCA_ctrl$v,
  colnames(Ya),
  component = 1,
  top_n = top_n_heatmap
)

mat_mtb_ctrl <- make_temporal_matrix(
  Xa,
  meta_a,
  names(mtb_loadings_ctrl)
)

mat_mb_ctrl <- make_temporal_matrix(
  Ya,
  meta_a,
  names(mb_loadings_ctrl)
)

mtb_labels_ctrl <- wrap_labels(
  rownames(mat_mtb_ctrl),
  width = 28
)

mb_labels_ctrl <- wrap_labels(
  rownames(mat_mb_ctrl),
  width = 28
)

ph_mtb_ctrl <- pheatmap(
  mat_mtb_ctrl,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  display_numbers = FALSE,
  color = colorRampPalette(c("white", "navy"))(50),
  labels_row = mtb_labels_ctrl,
  main = "Control C1 – Metabolome",
  fontsize_row = 11,
  fontsize_col = 10,
  angle_col = 0,
  border_color = NA,
  legend = TRUE,
  silent = TRUE
)

ph_mb_ctrl <- pheatmap(
  mat_mb_ctrl,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  display_numbers = FALSE,
  color = colorRampPalette(c("white", "firebrick3"))(50),
  labels_row = mb_labels_ctrl,
  main = "Control C1 – Microbiome",
  fontsize_row = 11,
  fontsize_col = 10,
  angle_col = 0,
  border_color = NA,
  legend = TRUE,
  silent = TRUE
)

combined_heatmap_ctrl <- arrangeGrob(
  ph_mtb_ctrl$gtable,
  ph_mb_ctrl$gtable,
  ncol = 2
)

grid.newpage()
grid.draw(combined_heatmap_ctrl)

# control_file <- file.path(
#   "../inst/SavedFigures",
#   "Chicken_control_feature_heatmaps.pdf"
# )
#
# pdf(control_file, width = 14, height = 6.5)
# grid.newpage()
# grid.draw(combined_heatmap_ctrl)
# dev.off()
```

### G. Why use a multimodal analysis?

A unimodal analysis can identify strong variation within the metabolome or microbiome, but it does not require the same sample groups to differ coherently in both data types. In contrast, s3CA estimates paired loading vectors and prioritizes structures for which the metabolome and microbiome scores vary together after ancillary adjustment.

The following diagnostic compares the fraction of score variation attributable to the displayed time groups in:

- the metabolome score alone;
- the microbiome score alone;
- the joint two-dimensional score pair.

This is a descriptive summary, not a formal proof that multimodal analysis is universally superior.

```{r multimodal_justification, eval = TRUE, fig.width=15, fig.height=5, out.width="100%"}
## Fit contrastive PCA separately to each modality
cpca_mtb <- fit_cpca(
  target_data = Xt,
  ancillary_data = Xa,
  n_components = 2
)

cpca_mb <- fit_cpca(
  target_data = Yt,
  ancillary_data = Ya,
  n_components = 2
)

cat(sprintf(
  "Metabolome cPCA scaling parameter: %.6f\n",
  cpca_mtb$alpha
))

cat(sprintf(
  "Microbiome cPCA scaling parameter: %.6f\n",
  cpca_mb$alpha
))

cat(
  "Metabolome contrastive eigenvalues:",
  paste(round(cpca_mtb$values, 4), collapse = ", "),
  "\n"
)

cat(
  "Microbiome contrastive eigenvalues:",
  paste(round(cpca_mb$values, 4), collapse = ", "),
  "\n"
)


df_cpca_mtb <- data.frame(
  cPC1 = cpca_mtb$scores[, 1],
  cPC2 = cpca_mtb$scores[, 2],
  meta_t,
  check.names = FALSE
)

df_cpca_mb <- data.frame(
  cPC1 = cpca_mb$scores[, 1],
  cPC2 = cpca_mb$scores[, 2],
  meta_t,
  check.names = FALSE
)

df_cpca_mtb$Time_raw <- as.character(df_cpca_mtb$Time)
df_cpca_mb$Time_raw  <- as.character(df_cpca_mb$Time)

df_cpca_mtb$Time_plot <- make_time_label(
  df_cpca_mtb$Time,
  merge_late = merge_late
)

df_cpca_mb$Time_plot <- make_time_label(
  df_cpca_mb$Time,
  merge_late = merge_late
)


p_cpca_mtb <- plot_score_pair(
  df_cpca_mtb,
  x = "cPC1",
  y = "cPC2",
  title = "Metabolome-only cPCA",
  x_label = "Contrastive PC1",
  y_label = "Contrastive PC2",
  show_centroid_labels = FALSE
)

p_cpca_mb <- plot_score_pair(
  df_cpca_mb,
  x = "cPC1",
  y = "cPC2",
  title = "Microbiome-only cPCA",
  x_label = "Contrastive PC1",
  y_label = "Contrastive PC2",
  show_centroid_labels = FALSE
)

p_s3ca_joint <- plot_score_pair(
  df_ctst_c1,
  x = "cCCX1",
  y = "cCCY1",
  title = "Joint microbiome–metabolome s3CA",
  x_label = "Contrastive metabolome score",
  y_label = "Contrastive microbiome score",
  show_centroid_labels = FALSE
)

p_multimodal_comparison <- ggarrange(
  p_cpca_mtb,
  p_cpca_mb,
  p_s3ca_joint,
  nrow = 1,
  common.legend = TRUE,
  legend = "bottom"
)

print(p_multimodal_comparison)

# multimodal_file <- file.path(
#   "../inst/SavedFigures",
#   if (merge_late) {
#     "Chicken_unimodal_cpca_vs_s3ca_late_merged.pdf"
#   } else {
#     "Chicken_unimodal_cpca_vs_s3ca.pdf"
#   }
# )

# ggsave(
#   filename = multimodal_file,
#   plot = p_multimodal_comparison,
#   width = 15,
#   height = 5.2
# )

# cat(
#   "Saved unimodal cPCA versus multimodal s3CA comparison to:",
#   multimodal_file,
#   "\n"
# )
```


